Multiple dynamic interactions from basal ganglia direct and indirect pathways mediate action selection

Version 1

Abstract

Summary

The basal ganglia are known to be essential for action selection. However, the functional role of basal ganglia direct and indirect pathways in action selection remains unresolved. Here by employing cell-type-specific neuronal recording and manipulation in mice trained in a choice task, we demonstrate that multiple dynamic interactions from the direct and indirect pathways control the action selection. While the direct pathway regulates the behavioral choice in a linear manner, the indirect pathway exerts a nonlinear inverted-U-shaped control over action selection, depending on the inputs and the network state. We propose a new center (direct) - surround (indirect) - context (indirect) “Triple- control” functional model of basal ganglia, which can replicate the physiological and behavioral experimental observations that cannot be simply explained by either the traditional “Go/No-go” or more recent “Co-activation” model. These findings have important implications on understanding the basal ganglia circuitry and action selection in health and disease.

Introduction

Selecting the proper actions is essential for organism’s survival and reproduction in the ever-changing environment (Gallistel, 1980). Numerous studies have implicated that the basal ganglia, a series of interconnected subcortical nuclei including the striatum and substantia nigra, play a primary role in action selection (Graybiel, 1998; Hikosaka et al., 1998; Jin and Costa, 2015; Mink, 2003; Redgrave et al., 1999). Indeed, a wide range of neurological and psychiatric disorders associated with the dysfunctional basal ganglia circuitry, including Parkinson’s disease (Benecke et al., 1987), Huntington’s disease (Phillips et al., 1995), Obsessive-compulsive disorder (OCD) (Graybiel and Rauch, 2000), are characterized by major deficits in action selection and movement control. Anatomically, commands for motor control are processed by basal ganglia through two major pathways, termed direct and indirect pathway, originating from striatal D1- and D2-expressing spiny projection neurons (D1-/D2-SPNs), respectively (Albin et al., 1989; DeLong, 1990). These two pathways collectively modulate substantia nigra pars reticulata (SNr) activity and the basal ganglia output, thus influence behavioral decisions. There are currently two major types of thinking on how the basal ganglia pathways work. An early classic theory has suggested that the basal ganglia direct and indirect pathways oppose each other to facilitate and inhibit action, respectively (the “Go/No-go” model) (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). In contrast, a recent theory has proposed that direct pathway selects the desired action, while the indirect pathway inhibits other competing actions in order to highlight the targeted choice (the “Co-activation” model) (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996).

The two theories have essentially agreed upon the function of direct pathway being the positive driving force for initiating or facilitating the desired actions. Yet, the ideas about the indirect pathway function are largely controversial as either impeding the desired action in the “Go/No-go” model or inhibiting the competing actions in the “Co-activation” model. While the precise neuroanatomy on how the D2-SPNs control SNr through indirect pathway has yet to be mapped out at single-cell level to differentiate the two hypotheses, either theory has found its supports from behavioral and physiological observations. For instance, it has been found that stimulation of striatal direct and indirect pathways can bidirectionally regulates locomotion (Durieux et al., 2012; Kravitz et al., 2010), consistent with the traditional ‘Go/No-go’ model. On the other hand, in vivo electrophysiological and imaging experiments revealed that the striatal direct and indirect pathways are both activated during action initiation (Barbera et al., 2016; Cui et al., 2013; Geddes et al., 2018; Isomura et al., 2013; Jin et al., 2014; Klaus and Plenz, 2016; Markowitz et al., 2018; Nonomura et al., 2018), as the ‘Co-activation’ model predicted. Furthermore, physiological and optogenetic studies concerning complex behavior such as learned action sequences have further complicated the issue, and unveiled various neuronal subpopulations in both pathways are activated during the initiation, termination and switching of actions (Geddes et al., 2018; Jin and Costa, 2015; Jin et al., 2014; Tecuapetla et al., 2016). So far, how exactly the basal ganglia direct and indirect pathways work together to control action selection has been controversial and inconclusive, and the underlying circuit mechanism remains largely unclear (Calabresi et al., 2014).

Here we trained mice to perform an operant action selection task where they were required to select one out of two actions to achieve reward, based on self-monitored time intervals (Howard et al., 2017). By employing in vivo neuronal recording, we found that the net output of two opponent SNr neuron populations is predictive of the behavioral choices. Through identifying striatal pathway-specific neuronal activity with optogenetic tagging, we found that there are neuronal subpopulations in either direct or indirect pathway who are activated during selecting one action and suppressed during another. Optogenetic inhibition, as well as selective ablation of direct pathway impairs action selection, and optogenetic excitation of direct pathway enhances current choice, confirming a role of direct pathway in facilitating desired actions. Furthermore, optogenetic inhibition of indirect pathway improves action selection and excitation of indirect pathway impairs behavioral choices, as predicted by the ‘Go/No-go’ model. However, selective ablation of indirect pathway impairs action selection, opposite from the behavioral effect of optogenetic inhibition and at odds with the ‘Go/No-go’ model, but in consistent with the prediction from the ‘Co-activation’ model. To resolve these contradictions, we propose a new center (direct) - surround (indirect) - context (indirect) “Triple-control” functional model of basal ganglia pathways, in which there are two interacting indirect pathway subcircuits exerting opposite controls over the basal ganglia output. The new model can reproduce the neuronal and behavioral experimental results that cannot be simply explained by either the “Go/No-go” or the “Co-activation” model. Further systematic analyses from this new model suggested that the direct and indirect pathways modulate behavioral outputs in a linear and nonlinear manner, respectively. Notably, in the new ‘Triple-control’ model, the direct and indirect pathways can work together to dynamically control action selection and operate in a manner similar to ‘Go/No-go’ or ‘Co-activation’ model, depending on the activity level and the network state. These results revise our current understanding on how the basal ganglia control actions, and have important implications for a wide range of movement and psychiatric diseases where the dynamic balance between the two pathways is compromised (Albin et al., 1989; Benecke et al., 1987; Calabresi et al., 2014; DeLong, 1990; Graybiel and Rauch, 2000; Mink, 1996; Phillips et al., 1995).

Results

Opponent SNr activities underlie action selection

To address the role of basal ganglia in action selection, we trained mice in a recently developed 2-8 s task in which they are required to choose the left versus right action based on self-monitored time intervals (Howard et al., 2017). Specifically, mice were put into an operant chamber with both left and right levers extended (Figure 1A, see Methods). For a given trial, both levers retract at trial initiation, and after either 2 s or 8 s (50% for each, randomly interleaved), both levers extend. The mouse has to judge the interval between lever retraction and extension as 2 s vs. 8 s and make a corresponding action choice by pressing the left vs. right lever, respectively (Figure 1A). The first lever press after lever extension was registered as the mouse’s choice. The correct choice leads to sucrose delivery (10 µl) as reward, and any lever presses beyond the first press after lever extension yield no outcome. A new trial starts at lever retraction again after a random inter-trial-interval (ITI, 30 s on average; Figure 1A). Across 14 consecutive days of training, mice (n = 10) significantly increased the correct rate of choice from chance level to more than 90% (Figure 1B). In addition, the animals gradually shortened the choice latency and demonstrated a strong preference toward the left lever due to its association with the shorter waiting time (Figure S1A, B). As a result, during the longer-waiting 8 s trials the mouse initially moved toward the left lever, then crossing the midpoint between left and right levers at around 4 s, and stayed around the right lever afterwards (Howard et al., 2017) (Figure 1C; Supplemental Video 1). Note that, the mouse showed no stereotyped movement trajectories during the incorrect trials (Figure 1C). This emerged stereotyped movement trajectory in the 8-s trials thus provided us a unique opportunity for investigating the neural mechanisms underlying the internally-driven, dynamic action selection process.

The neuronal dynamics in SNr during the 2-8 s action selection task.

(A) Schematic diagram for the design of 2-8 s task. (B) Correct rate for wild type mice across 14 days training (n = 10 mice, one-way repeated-measures ANOVA, significant effect of training days, F13,117 = 32.54, p < 0.0001). (C) Movement trajectory of an example mouse in correct (left panel) and incorrect (right panel) 8-s trials (gray line: trajectory of each trials; red/black line: the average trajectory). (D) Diagram of electrode array implanted into substantia nigra pars reticulata (SNr). (E) Firing Rate Index (FRI) of neuronal activity for all task-related SNr neurons in correct 8-s trials. The magnitude of FRI is color coded and the SNr neurons are classified as four different types based on the activity dynamics. (F-I) Averaged FRI for Type 1 (F), Type 2 (G), Type 3 (H), Type 4 (I) of SNr neurons in correct (red) and incorrect 8-s trials (gray). (J) The proportion of four types of SNr neurons. Type 1 and Type 2 are major types and significantly more than Type 3 and Type 4 (Z-test, p < 0.05). (K) Integrated SNr output defined as the subtraction of averaged FRI between Type 1 and Type 2 SNr neurons. (L) Averaged psychometric curve (n = 10 mice) of choice behavior. (M) The correlation between the Type 1 and Type 2 FRI subtraction and the behavioral choice (R = 0.98, p < 0.0005). Error bars denote s.e.m., same for below unless stated otherwise.

The substantia nigra pars reticulata (SNr) is one of the major output nuclei of basal ganglia (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). To investigate how the basal ganglia contribute to the dynamic process of action selection, we began by recording the SNr neuronal activity in mice trained in the 2-8 s task (Figure 1D, Figure S1C, see Methods). It was found that a large proportion (211/261, 80.8%; recorded from n = 9 mice) of SNr neurons changed firing rate significantly during the correct 8-s trials as mice dynamically shifted the internal action selection from the left to the right (Figure 1E). The Z-score of the task-related neuronal firing rate, reflecting the firing activity changes related to baseline, was defined as Firing Rate Index (FRI, see Methods). We focus on the data analyses in the 8-s trials since the first 2-s of 8-s trials consists of the identical behavioral and neuronal profiles of the 2-s trials due to the task design (Figure 1C, Figure S1D-F). The task-related SNr neurons were categorized into four subtypes based on the dynamics of FRI in the correct 8-s trials: Type 1 - monotonic decrease (Figure 1E, F, 102/211, 48.3%), Type 2 - monotonic increase (Figure 1E, G, 56/211, 26.5%), Type 3 - transient phasic increase (Figure 1E, H, 25/211, 11.9%) and Type 4 - transient phasic decrease (Figure 1E, I, 28/211, 13.3%). These types of neuronal dynamics in SNr only appeared in the correct but not the incorrect trials (Figure 1F-I), nor on the day 1 of task training (Figure S1G-K), suggesting a tight correlation between the SNr neuronal dynamics and the behavioral performance. Specifically, the Type 1 and Type 2, but not the Type 3 and Type 4 neurons, exhibit firing changes co-varying with the action selection and these two types together consist in around 80% of all task-related SNr neuron population (Figure 1J).

It has been suggested that SNr suppresses movements through the inhibition of downstream motor nuclei and releases action via disinhibition (Hikosaka et al., 2000; Mink, 1996). We thus ask whether the opponent neuronal dynamics in Type 1 and Type 2 SNr subpopulations mediate the dynamic shift of choice, by suppressing the competing selection of right vs. left action, respectively. Indeed, the SNr net output by subtracting Type 2 and Type 1 SNr neuronal dynamics (Figure 1K) is highly reminiscent of the animal’s stereotyped movement trajectory during choice (Figure 1C). To further determine the relationship between the SNr net output and action selection, we tested the behavioral choice of the 2-8 s trained mice in a series of non-rewarded probe trials with novel intervals of 2.5, 3.2, 4, 5 and 6.3 s (see Methods). Consistent with what reported before (Howard et al., 2017), the probability of mice selecting the action associated with the long duration (8 s) gradually increases along with the time intervals of probe trials (Figure 1L). The resulting psychometric curve thus represents the animal’s real-time action selection process during the 8-s trials. Further comparison between the psychometric curve and the SNr net output revealed a strong linear correlation (Figure 1M), indicating that the SNr net output faithfully predicts momentary behavioral choice. Together, these results suggest that mice can learn to dynamically shift their choice based on internally-monitored time, and the opponent neuronal activities in SNr correlate with the action selection.

SNr neuronal dynamics reflect action selection but not simply time or value

In the 2-8 s task, the passage of time and expectation of reward both change simultaneously with the animal’s internal choice. One may argue that the Type 1 and Type 2 neuronal dynamics observed in SNr during the 8-s trials might reflect the passage of time or value of expected reward rather than action selection. To differentiate these possibilities and specify the functional role of SNr activity, we presented mice previously trained in the 2-8 s task with random probe trials of 16-s interval (Figure 2A). In these 16-s probe trials which they have never experienced before during training, the animals sometimes wait on the right side and press the right lever, or shift back to the left side and press the left lever when the levers are extended at 16 s (Figure 2B, C). This arbitrary choice situation in the 16-s probe trials thus provides a special window to determine the functional relationship between SNr activity and behavioral choice. If the Type 1 and Type 2 SNr subpopulations encode information about time passage or expectation value, their neuronal activities would continue changing monotonically between 8 and 16 s. In contrast, if the Type 1 and Type 2 SNr subpopulations encode action selection, their neuronal activities would predict the behavioral choice and differentiate between the right vs. left action selection. Indeed, it was found that when the firing activity of Type 1 SNr neurons maintained below baseline from 8 to 16 s, the mice tended to select the right lever later (Figure 2D). However, when the firing activity reversed the decreasing tendency to increase, the mice chose the left lever instead (Figure 2D). A similar relationship between the neuronal activity and behavior choice was also evident in Type 2 SNr neurons, albeit with opposite dynamics (Figure 2E). This is especially evident in the subtraction of Type 2 and Type 1 SNr neuronal dynamics, in which the SNr net output is strongly correlated with and predictive of behavioral choice (Figure 2F). These results thus suggested that the neuronal activities in SNr likely encode the ongoing action selection but not simply reflect time passage or reward value.

SNr neuronal dynamics reflect action selection but not interval time or reward value.

(A) Task diagram of 2-8 s control task with 10% 16-s probe trials. (B) Percentage of behavioral choice in 2-s, 8-s and 16-s trials (blue: left choice; red: right choice) (n = 9 mice, paired t-test, p < 0.05). (C) Movement trajectory of an example mouse in 16-s trials (blue: left choice; red: right choice). (D) Averaged SNr Type 1 FRI in 16-s trials (red: left choice; black: right choice). Firing rates from 8s to 16s (highlighted area) are compared between left and right choice (n = 26 neurons, two-way repeated-measures ANOVA, significant difference between left and right choices, F1,25 = 6.646, p = 0.016). (E) Averaged SNr Type 2 FRI in 16-s trials (red: left choice; black: right choice). Firing rates from 8s to 16s are compared between left and right choice (n = 16 neurons, two-way repeated-measures ANOVA, significant difference between left and right choices, F1,15 = 5.785, p = 0.029). (F) Subtraction of FRI for SNr Type 1 and Type 2 neurons in 16-s probe trials (red: left choice; black: right choice). (G) Task design of 2-8 s standard task. (H) Percentage of behavioral choice in 2-s and 8-s trials (blue: left choice; red: right choice) (n = 6 mice, paired t-test, p < 0.05). (I) Movement trajectory of an example mouse in 8-s trials (blue: left choice; red: right choice). (J) Task design of reversed 2-8 s task. (K) Percentage of behavioral choice in 2-s and 8-s trials in the reversed 2-8 s task (blue: left choice; red: right choice) (n = 6 mice, paired t-test, p < 0.05). (L) Movement trajectory of the same mouse as (I) in 8-s trials in the reversed 2-8 s task (blue: left choice; red: right choice). (M) Averaged FRI of the SNr Type 1 neurons in correct 8-s trials (n = 14 neurons). (N) Averaged FRI of the SNr Type 2 neurons in correct 8-s trials (n = 11 neurons). (O) Integrated SNr output as the subtraction of FRI for SNr Type 1 (M) and Type 2 neurons (N) in the standard 2-8 s task. (P) Averaged FRI of the same neurons as (M) in correct 8-s trials of the reversed 2-8 s task. (Q) Averaged FRI of the same neurons as (N) in correct 8-s trials of the reversed 2-8 s task. (R) Integrated SNr output as the subtraction of FRI for SNr Type 1 (P) and Type 2 neurons (Q) in the reversed 2-8 s task.

To further confirm this point, we recorded the firing activity from the same SNr neurons during both the 2-8 s control task (Figure 2G, 2s-left and 8s-right) and a modified version of 2-8 s task in which the contingency between action and interval is reversed (Figure 2J, 2s-right and 8s-left) on the same day (see Methods). It was found that the mice performed at around 80% correct in both tasks on the same day (Figure 2H, K, Figure S2A). Accordingly, the movement trajectories of the same mice in 8-s trials were reversed from left-then-right in the control task (Figure 2I) to right-then-left in the reversed task (Figure 2L). The left-lever preference during the ITI in the control task was also switched to right-lever preference in the reversed 2-8 s task (Figure S2B). Notably, the passage of time and expected value as well as other environmental factors are all identical in both versions of task, except that the animal’s choice is now reversed from right to left for the 8-s trials (Figure 2H, I vs. Figure 2K, L). If Type 1 or Type 2 SNr neurons encode time or value, either neuronal population will exhibit the same neuronal dynamics in 8-s trials for both versions of task. On the other hand, if Type 1 and Type 2 SNr neurons encode action selection, their neuronal dynamics will reverse in the reversed version of 2-8 s task compared to the standard version. In fact, the Type 1 SNr neurons which showed monotonic decreasing dynamics in the control 2-8 s task (Figure 2M) reversed their neuronal dynamics to a monotonic increase in the reversed 2-8 s task (Figure 2P), consistent with the behavioral choice. The same reversal of neuronal dynamics was also observed in Type 2 SNr neurons in the reversed version of standard task (Figure 2N, Q). The SNr net output by subtracting Type 2 and Type 1 SNr neuronal dynamics, which was tightly correlated with the action selection in the standard 2-8 s task (Figure 2O), is reversed and now predictive of the new behavioral choice in the reversed 2-8 s task (Figure 2R). Notably, Type 3 and Type 4 SNr neurons exhibiting transient change when mice switching between choices maintained the same neuronal dynamics in both tasks (Figure S2C-F). Together, these results therefore demonstrate that the output of basal ganglia reflects the dynamic action selection rather than simply time or value.

Distinct striatal direct vs. indirect pathway activity during action selection

The basal ganglia output is largely controlled by two major neural pathways, called ‘direct’ and ‘indirect’ pathway, originating from D1- vs. D2-expressing spiny projection neurons (D1- vs. D2-SPNs) in the striatum, respectively (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). We then decided to determine the neuronal dynamics in the striatum, specifically the neuronal activity in the direct and indirect pathways during action selection. We employed in vivo extracellular electrophysiology to record the neuronal activity in the dorsal striatum when mice perform the 2-8 s task, and classified putative SPNs based on the spike waveforms and firing properties (Geddes et al., 2018; Jin and Costa, 2010; Jin et al., 2014). Among all the SPNs recorded from the trained mice (n = 24), 341 out of 409 SPNs (83.4%) were defined as task-related neurons for showing significant firing changes during the 2-s and 8-s lever retraction period (Figure 3A, Figure S3A). Similar to the various types of neuronal dynamics observed in SNr, task-related SPNs showed Type 1 (Figure 3B, monotonic decrease, 159/341, 46.6%), Type 2 (Figure 3C, monotonic increase, 103/341, 30.2%), Type 3 (Figure 3D, transient phasic increase, 49/341, 14.4%) and Type 4 (Figure 3E, transient phasic decrease, 30/341, 8.8%) activity profiles during the correct 8-s trials (Figure 3A, Figure S3A). These neural dynamics were largely absent in SPNs on day 1 of training (Figure S3B-F). These results indicate that the striatum, as one of the major input nuclei of basal ganglia, demonstrates the four types of neuronal dynamics similar with SNr during the dynamic process of action selection.

Neuronal activity of striatal D1- and D2-SPNs during action selection.

(A) FRI of neuronal activity for all task-related SPNs in correct 8-s trials. SPNs were classified as Type 1 - 4. (B-E) Averaged FRI for Type 1 (B), Type 2 (C), Type 3 (D), Type 4 (E) of SPNs in correct (red) and incorrect 8-s trials (gray). (F) Diagram of simultaneous neuronal recording and optogenetic identification of D1- vs. D2-SPNs in dorsal striatum. (G) Top panel: Raster plot for a representative D1-SPN response to 100 ms optogenetic stimulation. Each row represents one trial and each black dot represents a spike. Bottom panel: Peristimulus time histogram (PETH) aligned to light onset at time zero. (H) PETH for the same neuron as shown in (G) with a finer time scale. (I) Distribution of light response latencies for D1- and D2-SPNs. (J) Action potential waveforms of the same neuron in (G) for spontaneous (black) and light-evoked (orange) spikes (R = 0.998, P < 0.0001, Pearson’s correlation). (K) Principal component analysis (PCA) of action potential waveforms showing the overlapped clusters of spontaneous (black) and light-evoked (orange) spikes. (L) Proportion of D1-SPN subtypes. Type 1 neurons are significantly more than other three types of neurons in D1-SPNs (Z-test, p < 0.05). (M) Averaged FRI for Type 1 (blue) and Type 2 (red) D1-SPNs in correct 8-s trials. (N) Proportion of D2-SPN subtypes. (O) Averaged FRI for Type 1 (blue) and Type 2 (red) D2-SPNs in correct 8-s trials.

To further determine the neuronal activity in the direct and indirect pathways during action selection, we utilized an optogenetics-aided photo-tagging method (Geddes et al., 2018; Howard et al., 2017; Jin and Costa, 2010; Jin et al., 2014; Lima et al., 2009) to record and identify striatal D1- vs. D2-SPNs in freely behaving mice. Channelrhodopsin-2 (ChR2) was selectively expressed in D1- or D2-SPNs by injecting AAV-FLEX-ChR2 in the dorsal striatum of D1- and A2a-Cre mice, respectively (Geddes et al., 2018; Jin et al., 2014). In the end of each behavioral session with recording, optogenetic stimulation via an optic fiber attached to the electrode array was delivered to identify D1- vs. D2-SPNs through photo-tagging (Figure 3F, Figure S3G-J) (Geddes et al., 2018; Jin et al., 2014). Only those neurons exhibiting a very short latency (≤ 6 ms) to light stimulation (Figure 3G-I) and showing identical spike waveforms (R ≥ 0.95, Pearson correlation coefficient) between behavior and light-evoked response (Figure 3J, K) were identified as Cre-positive thus D1- or D2-SPNs (Geddes et al., 2018; Jin et al., 2014). Within all positively identified D1-SPNs (n = 92 from 6 mice) and D2-SPNs (n = 95 from 6 mice), 74 out of 92 (80.4%) D1-SPNs and 79 out of 95 (83.1%) D2-SPNs showed a significant change in firing rate during the correct 8-s trials. In addition, all four types of neuronal dynamics during action selection were found in both D1-SPNs (Figure 3L, M) and D2-SPNs (Figure 3N, O), as observed in SNr. The Type 1 and Type 2 neuronal dynamics showing monotonic firing change (Figure 3M, O) were the predominant task-related subpopulations within either D1- (Figure 3L) or D2-SPNs (Figure 3N). Notably, the striatal D1-SPNs consist of significantly more Type 1 than Type 2 neurons (Figure 3L), while D2-SPNs show a similar proportion between the two Types (Figure 3N). These data thus suggest while neurons in both the striatal direct and indirect pathways encode information related to behavioral choice, the two pathways might reflect and contribute to distinct aspects of action selection.

Ablation of striatal direct vs. indirect pathway differently impaired action selection

Given the action-selection-related neuronal dynamics observed in striatum, we next asked whether the neural activity in striatum is necessary for learning and execution of action selection, and furthermore, what is the functional difference between the direct and indirect pathways. It has been reported that the NMDA receptors on striatal SPNs are critical for sequence learning (Geddes et al., 2018; Jin and Costa, 2010) and action selection (Howard et al., 2017). To further identify the functional role of NMDA receptors on D1- vs. D2-SPNs for action selection, we employed a genetic strategy to specifically delete NMDA receptors from D1- vs. D2-SPNs by crossing mice carrying a floxed NMDAR1 (NR1) allele with a dorsal striatum-dominant D1-cre line (Gong et al., 2007) and A2a-cre line (Geddes et al., 2018; Jin et al., 2014), respectively (referred to as D1-NR1 KO and D2-NR1 KO mice, respectively; see Methods). Both the D1-NR1 KO and D2-NR1 KO mice are significantly impaired in learning the 2-8 s task compared to their littermate controls (Figure 4A, B), suggesting that NMDA receptors on either D1- or D2-SPNs are critical for learning of proper action selection. In the end of two-week training, when given the probe trials with various intervals across 2 to 8 s, it was found that D1-NR1 KO mice showed a systematic bias toward the lever associated with short interval and made deficient behavioral choice only in long interval trials (Figure 4C). In contrast, D2-NR1 KO mice showed impaired action selection across various probe trials of both short and long intervals (Figure 4D). These data suggest that while NMDA receptors on both D1- and D2-SPNs are required for action learning, the deletion of NMDA receptors in direct and indirect pathways impairs action selection in a different manner.

Selective genetic knockout and ablation of D1- or D2-SPNs distinctly alters action selection.

(A) Correct rate of control (n = 11 mice) and D1-NR1 KO mice (n = 16) in 2-8 s task during 14 days training (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 10.8, p = 0.003). (B) Correct rate of control (n = 17) and D2-NR1 KO mice (n = 10) in 2-8 s task during 14 days training (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 8.728, p = 0.007). (C) The psychometric curve for control (n = 11) and D1-NR1 KO mice (n = 16) (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 12.27, p = 0.002). (D) The psychometric curve for control (n = 17) and D2-NR1 KO mice (n = 10) (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 9.64, p = 0.005). (E) Schematic of muscimol infusion into the dorsal striatum in trained mice. (F) Correct rate for control (black: pre-muscimol, gray: post-muscimol) and mice with muscimol infusion (magenta) in dorsal striatum (n = 9 mice, paired t-test, p < 0.01). (G) The psychometric curve for control (n = 9 mice, black: pre-muscimol, gray: post-muscimol control) and mice with muscimol infusion (n = 9 mice, magenta) in dorsal striatum (two-way repeated-measures ANOVA, significant difference between control and muscimol infusion, F2,16 = 11.74, p = 0.0007). (H) Timeline for selective diphtheria toxin (DT) ablation experiments. (I) Schematic of diphtheria toxin receptor (DTR) virus (AAV-FLEX-DTR-GFP) injection in dorsal striatum of D1-Cre mice. (J) Correct rate for control (n = 9 mice) and mice with dorsal striatum D1-SPNs ablation (D1-DTR, n = 8 mice) (two-sample t-test, p = 0.0016). (K) The psychometric curve for control (n = 9 mice) and D1-SPNs ablation mice (n = 8 mice) (two-way repeated-measures ANOVA, main effect of ablation, F1,15 = 1.84, p = 0.195; interaction between trial intervals and ablation, F6,90 = 4.14, p = 0.001). (L) Movement trajectory of a control mouse in 8-s trials. (M) Movement trajectory of a D1-DTR mouse in 8-s trials. (N) Schematic of diphtheria toxin receptor (DTR) virus (AAV-FLEX-DTR-GFP) injection in dorsal striatum of A2a-Cre mice. (O) Correct rate for control (n = 8 mice) and mice with dorsal striatum D2-SPNs ablation (D2-DTR, n = 8 mice) (two-sample t-test, p = 0.005). (P) The psychometric curve for control (n = 9 mice) and D2-SPNs ablation mice (n = 8 mice) (two-way repeated-measures ANOVA, main effect of ablation, F1,15 = 0.477, p = 0.5; interaction between trial intervals and ablation, F6,90 = 12.6, p < 0.001). (Q) Movement trajectory of a control mouse in 8-s trials. (R) Movement trajectory of a D2-DTR mouse in 8-s trials. (S) Schematic of center-surround receptive field diagram for Go/No-Go (left) and Co-activation (right) models. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection.

We then asked whether that neural activity in dorsal striatum is necessary for the proper execution of action selection after learning. We first conduct striatal inactivation in trained wildtype mice by bilateral intra-striatal infusion of muscimol (Figure 4E, see Methods). Striatal muscimol infusion significantly reduced the animal’s overall performance in comparison with the pre- and post-saline injection controls (Figure 4F). When tested with probe trials, the psychometric curve indicated that the inactivation of striatum impairs action selection for the long trials (Figure 4G). These data thus suggested that the striatal neural activity is critical for appropriate execution of learned action selection.

To further elucidate the functional role of specific striatal pathways in action selection, we next employed a viral approach to bilaterally express diphtheria toxin receptors (AAV-FLEX-DTR-eGFP) in the dorsal striatum of trained D1- and A2a-Cre mice, followed by diphtheria toxin (DT) injections to selectively ablate D1- or D2-SPNs (Geddes et al., 2018) (Figure 4H, I, N; see Methods). Ablation of either D1- or D2-SPNs significantly impaired action selection and reduced the correct rate of choice (Figure 4J, O). Notably, the psychometric curve revealed that D1-SPNs ablation mice showed a selective impairment of choice in long interval trials (Figure 4K). In contrast, mice with D2-SPNs ablation exhibited choice deficits in both long and short trials (Figure 4P). Consistent with the D1- and A2a-NR1 KO data, these results suggest that the direct and indirect pathways are both needed yet play distinct roles in action selection.

The classic ‘Go/No-go’ model of basal ganglia suggests the direct and indirect pathways work antagonistically to release and inhibit action, respectively (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). On the other hand, more recent ‘Co-activation’ model of basal ganglia proposes that direct pathway initiates the selected action and at the same time, the indirect pathway inhibits the competing actions (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). For visualization purpose, we diagram ‘Go/No-go’ and ‘Co-activation’ models as center-surround receptive field with D1-SPNs as the center and D2-SPNs as the surround (Figure 4S; Figure S4A, D). While the direct pathway plays the similar role in both models (Figure S4B, E), the function of indirect pathway differs dramatically (Figure 4S). Lesion of the indirect pathway thus leads to contrast predictions on action selection from the two models (Figure S4C, F). Specifically, ablation of D2-SPNs would facilitate the action being selected through removing inhibition according to the Go/No-go model (Figure S4C) (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010), while blockage of indirect pathway would impair the action selection due to disinhibition of competing actions according to the Co-activation model (Figure S4F) (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). Although our D1-SPNs ablation experiment indicates that direct pathway is required for action selection as suggested in both models (Figure 4J, K), the D2-SPNs ablation result favorably supports the Co-activation model over the Go/No-go model (Figure 4O, P, Figure S4F). In fact, close inspection of the movement trajectories of D1-SPNs lesioned mice in the 8-s trials showed that compared to control mice (Figure 4L), they tend to stick on the left side more often with impaired right choice when lever extension at 8s (Figure 4M). In contrast, D2-SPNs lesioned mice demonstrated overall rather random movement trajectories, and the stereotyped left-then-right movement sequences were largely disrupted in comparison with the controls (Figure 4Q, R). These observations are mostly consistent with the idea of indirect pathway inhibiting competing actions in the Co-activation model (Figure 4S) and lesion of indirect pathway disrupts action selection for both the short and long trials (Figure 4P-R). Together, these data suggest that ablation of direct and indirect pathways both impair choice behavior but in a distinct manner due to their different roles in action selection.

Optogenetic manipulation of D1- vs. D2-SPNs distinctly regulates action selection

To further determine the specific function of direct vs. indirect pathway in action selection, we employed optogenetics to alter the D1- and D2-SPNs activity in vivo with high temporal precision and investigated its effects on the ongoing action selection process. Both the classic ‘Go/No-go’ (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010) and more recent ‘Co-activation’ (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996) models predict that activation of the direct pathway enhances the action selection (Figure S5A, E, I, K, O, Q), while inhibition of direct pathway reduces the correct choice (Figure S5B, F, L, R). To experimentally validate the models’ predictions, AAV-FLEX-ChR2 was injected into the dorsal striatum of D1- or A2a-Cre mice and optic fibers were implanted bilaterally for in vivo optogenetic stimulation (Figure 5A, Figure S3K, L; see Methods) (Geddes et al., 2018; Jin et al., 2014). After mice learned the 2-8s task, 1-s pulse of constant light (wave length 473 nm) was delivered right before lever extension in randomly chosen 50% of 2-s and 50% of 8-s trials (Figure 5B, C, see Methods). The correct rate of optogenetic stimulation trials is compared with the non-stimulation trials of the same animal as a within-subject design. We observed no significant change on the correct rate in 2-s trials, whereas the correct rate was significant increased by optogenetic stimulation in 8-s trials (Figure 5D), indicating a facilitation effect on action selection by stimulating the D1-SPNs. We then sought to determine the effect of suppressing D1-SPN activity on action selection by viral expression of Halorhodopsin (AAV5-EF1a-DIO-eNpHR3.0-eYFP) in the dorsal striatum of D1-cre mice (Gradinaru et al., 2010). As expected, inhibiting D1-SPNs right before lever extension in trained mice reduced the correct rates in 8-s but not 2-s trials (Figure 5C, E), opposite to D1-SPN stimulation effects. These experimental data with bidirectional optogenetic manipulation suggest that the D1-SPN activity is positively correlated with the choice performance, in consistent with the hypothesis of direct pathway facilitating the action selected in both the Go/No-go and Co-activation models (Figure S5K, L, Q, R).

Optogenetic manipulation of D1- vs. D2-SPNs differently regulates action selection.

(A) Schematic of optic fiber implantation for experimentally optogenetic excitation or inhibition of D1- or D2-SPNs in the dorsal striatum. (B, C) Schematic for optogenetic excitation (B) and inhibition (C) of D1-/D2-SPNs for 1 s right before lever extension in 2-8 s task. (D) Change of correct rate for optogenetic excitation of D1-SPNs in 2-s and 8-s trials (n = 11 mice, one-sample t-test, 2-s trials: p = 0.248; 8-s trials: p < 0.05). (E) Change of correct rate for optogenetic inhibition of D1-SPNs in 2-s and 8-s trials (n = 6 mice, one-sample t-test, 2-s trials: p = 0.557; 8-s trials: p < 0.05). (F) Change of correct rate for optogenetic excitation of D2-SPNs in 2-s and 8-s trials (n = 8 mice, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (G) Change of correct rate for optogenetic inhibition of D2-SPNs in 2-s and 8-s trials (n = 5 mice, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (H) Schematic of center-surround receptive field diagram for Go/No-Go (left) and Co-activation (right) models. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection.

Nevertheless, the two models have distinct views on the function of indirect pathway. While the classic ‘Go/No-go’ model suggests that the indirect pathway inhibits the selected action (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010), the ‘Co-activation’ model hypothesizes that the indirect pathway inhibits the competing actions instead (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). These models thus provide contrast predictions about the effect of activation of the indirect pathway on action selection, being decreased correct rate based on the Go/No-go model (Figure S5M) and increased correct rate from the Co-activation model (Figure S5S), respectively. We thus decided to test the distinct predictions from the two models by optogenetic manipulation of indirect pathway during action selection in the 2-8s task. ChR2 or Halorhodopsin (eNpHR3.0) was expressed in the dorsal striatum of A2a-cre mice for bilaterally optogenetic activation or inhibition during behavior (Figure 5A, Figure S3L; see Methods). Notably, optogenetic excitation of D2-SPNs for 1s right before lever extension decreased the correct rate in both 2-s and 8-s trials (Figure 5F). In contrast, transient optogenetic inhibition of D2-SPNs before behavioral choice increased correct rates for both 2-s and 8-s trials (Figure 5G). These data suggest that opposite to the D1-SPN manipulation, optogenetic stimulation of D2-SPNs impairs action selection, while inhibition of D2-SPNs facilitates behavioral choice. These optogenetic results further unveil the distinct roles of direct vs. indirect pathway in action selection, and are in line with the predictions from the Go/No-go (Figure 5H, Figure S5M, N) but not the Co-activation model (Figure S5S, T).

A ‘Triple-control’ model of basal ganglia circuit for action selection

Our DT lesion experiments found that ablation of indirect pathway impairs action selection (Figure 4O, P), as predicted from the Co-activation but not Go/No-go model (Figure 4S), while the optogenetic results suggested that inhibition of D2-SPNs enhances behavioral choice (Figure 5G), a result in favor of the Go/No-go rather than Co-activation model (Figure 5H). We wonder whether these seemly discrepant effects are attributed to a more complex circuit mechanism involving in the indirect pathway different from either the Go/No-go or Co-activation model. To systematically investigate the cell type- and pathway-specific mechanisms underlying action selection, we firstly add Go/No-go and Co-activation models together to examine the whether the resulted combination model could explain the experimental observations (Figure S4G). The lesion of D1-SPNs in the combination model indeed selectively impaired choice in long interval trials (Figure S4H). However, the effect of D2-SPNs ablation in the combination model was neutralized due to the opposing contributions from Go/No-go and Co-activation models respectively (Figure S4I). Based on these simulation results, none of the Go/No-go, Co-activation and combination models was able to fully capture the underlying mechanism of basal ganglia in action selection. Inspired by the data in current experiments, we decided to build a new computational model of the cortico-basal ganglia circuitry based on the realistic neuroanatomy (Aoki et al., 2019; Mailly et al., 2003; Schmidt and Berke, 2017; Taverna et al., 2008) and empirical neuronal physiology during action selection (Figures 1-3).

Different from the dual control of action by direct vs. indirect pathway in either the Go/No-go or Co-activation model (Figure S4), our new model adds an additional layer of control derived from the indirect pathway, thus called ‘Triple-control’ model for action selection. Specifically, the new model consists of one direct pathway and two indirect pathways defined as D2-SPN #1 and D2-SPN #2 two subpopulations, corresponding to the Co-activation and Go/No-go functional modules, respectively (Figure 6A, B). In addition, the indirect pathway D2-SPNs in the Co-activation module inhibits the indirect pathway D2-SPNs in the Go/No-go module through the well-known D2-SPN collaterals with the properties of short-term depression in the striatum (Gustafson et al., 2006; Schmidt and Berke, 2017; Taverna et al., 2008; Tecuapetla et al., 2007) (Figure 6A; see Methods), providing asymmetric modulation to D2-SPN subgroups and promoting Co-activation module as the dominant functional module at rest. In this ‘Triple-control’ basal ganglia model, striatal D1- and D2-SPNs associated with left and right actions receive excitatory inputs from corresponding cortical inputs (Figure 6A) to generate Type 1 and Type 2 neuronal dynamics (Figure S6A-D) (Lo and Wang, 2006). The D1- and D2-SPNs then regulate the SNr neuronal dynamics through the direct and indirect pathways, respectively (Figure S6) (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). The net SNr output (Figure S6F, I), which controls the downstream brainstem and thalamic circuits necessary for action selection (Hikosaka, 2007; Lo and Wang, 2006; Redgrave et al., 1999), will determine the final behavioral choice (Figure S6G, J). Our computational simulations showed that this ‘Triple-control’ network model could faithfully recapitulate the neuronal activity across the basal ganglia circuitry and predict the behavioral choice (Figure S6).

A triple-control computational model of basal ganglia direct and indirect pathways for action selection.

(A) Network structure of the cortico-basal ganglia model based on realistic anatomy and synaptic connectivity. (B) Schematic of center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (C) The psychometric curves of behavioral output in control (black) and D1-SPNs ablation condition (blue) in ‘Triple-control’ model (n = 10, two-way repeated-measures ANOVA, main effect of ablation, F1,18 = 98.72, p < 0.0001; interaction between trial intervals and ablation, F6,108 = 7.799, p < 0.0001). (D) The psychometric curves of behavioral output in control (black) and D2-SPNs ablation condition (red) in ‘Triple-control’ model (n = 10, two-way repeated-measures ANOVA, main effect of ablation, F1,18 = 99.54, p < 0.0001; interaction between trial intervals and ablation, F6,108 = 177.6, p < 0.0001). (E) Change of correct rate for optogenetic excitation of D1-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p = 0.407; 8-s trials: p < 0.05). (F) Change of correct rate for optogenetic excitation of D2-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (G) Change of correct rate for optogenetic inhibition of D1-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p = 0.28; 8-s trials: p < 0.05). (H) Change of correct rate for optogenetic inhibition of D2-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05).

To dissect the functional role of direct vs. indirect pathway in action selection, we simulate the cell ablation experiments and examine the behavioral output in the ‘Triple-control’ basal ganglia model. Ablation of D1-SPNs in the network model (Figure S6E) modulates both Type 1 and Type 2 SNr dynamics but in different magnitude due to the biased striatal input to SNr left output and mutual inhibition between SNr left vs. right outputs (Figure S6F; see Methods). As a result, the lesion causes a downward shift in the net SNr output, especially evident at the late section of 8 s (Figure S6G). This change in net SNr output predicts a behavioral bias towards left choice as seen in the psychometric curve (Figure 6C), in consistent with experimental results in mice with D1-SPNs ablation (Figure 4K). In contrast, ablation of D2-SPNs in the network model (Figure S6H), by removing the indirect pathways of both the Go/No-go and Co-activation modules, alters Type 1 and Type 2 SNr dynamics (Figure S6I) and change the net SNr output dramatically around 2s as well as 8s (Figure S6J). The model thus predicts behavioral choice deficits for both short and long trials during D2-SPNs ablation (Figure 6D), consistent with experimental observations (Figure 4P). Together, these data suggest that our new ‘Triple-control’ basal ganglia model, based on realistic neuroanatomy and empirical neuronal physiology, can perform action selection similar to the behavior of mice, and successfully replicate the pathway-specific lesion effects on choice.

We further simulate the neuronal and behavioral effects of optogenetic manipulation of D1- and D2-SPNs in the cortico-basal ganglia model. Consistent with the experimental results (Figure 5D, F), optogenetic stimulation of D1-SPNs facilitates the ongoing choice (Figure 6E), while optogenetic inhibition of D1-SPNs suppresses ongoing choice in the model (Figure 6G). In addition, optogenetic stimulation of D2-SPNs impairs the ongoing choice and causes switching (Figure 6F), while optogenetic inhibition of D2-SPNs facilitates ongoing choice, due to the now dominant Go/No-go module mediated by the short-term depression of D2 collaterals in the model (Figure 6H). Consistent with the experimental observations, the optogenetic inhibition effect is opposite from the D2-SPNs cell ablation in the model (Figure 6D).

We next investigate how the striatum influence SNr outputs. Since the collateral projection with STD in D2-SPNs is the key in our ‘Triple-control’ model to switch between Go/No-go and Co-activation modules, we first built a motif of indirect pathway with two D2-SPNs subgroups defined as D2-SPN #1 and D2-SPN #2 (Figure S7A). We tested this indirect pathway motif with monotonic neural dynamics observed in experiments meanwhile simulating the optogenetic activation at 1s and 7s (Figure S7D-I). The SNr therefore received more activation at 1s than at 7s (Figure S7J, K), suggesting that the D2-SPNs with short-term depression in collateral inhibition modulates SNr activities in a firing rate-dependent manner.

We next sought to test the model’s predictions and experimentally investigate the distinctions in modulating SNr activities between the direct and indirect pathways during action selection. In order to manipulate D1- or D2-SPNs and monitor SNr responses at the same time, we simultaneously implanted optogenetic fibers and recording array into striatum and SNr respectively on a single mouse (Figure S7L). While the mice performing the 2-8s task, optogenetic stimulation was delivered to activate either D1- or D2-SPNs. It was found that optogenetic activation of D1- or D2-SPNs caused both inhibition and excitation in Type 1 and Type 2 SNr neurons (Figure S7M-O). To further compare SNr activities responding to striatal activation at different time points during the lever retraction period, for a given trial, we activated D1-SPNs (or D2-SPNs) either at 1s or 7s after the lever retraction (Figure S7P-R). For direct pathway, the change of FRI in SNr activities caused by activation of D1-SPNs showed no significant difference between 1s and 7s (Figure S7P, S). For indirect pathway, activating D2-SPNs at 1s caused smaller activation of FRI than at 7s in Type 1 SNr neurons (Figure S7Q), whereas for Type 2 SNr neurons, activating D2-SPNs at 1s induced bigger FRI increase at 1s than at 7s (Figure S7R). Overall, activating D2-SPNs tended to bias the firing rate downward at 1s but upward at 7s in Type 1 SNr neurons, which was counteractive to the decreasing tendency of Type 1 SNr neuron (Figure S7T). In contrast, Type 2 SNr neurons showed higher FRI increase and smaller decrease in response to activating D2-SPNs at 1s than at 7s, which was opposing to the increasing dynamics of Type 2 SNr neurons (Figure S7T). This firing rate-dependent modulation on SNr activities through indirect pathway is consistent with the computational simulation (Figure S7J, K; Figure S8). Therefore, the underlying D2-SPNs collaterals might indeed be a key mechanism contributing to the modulation of SNr activity and action selection in vivo, as simulated in the ‘Triple-control’ model.

Taken together, our new ‘Triple-control’ basal ganglia model, based on realistic neuroanatomy and empirical neurophysiology, successfully reproduces both the lesion and optogenetic data we collected during the animal experiments. It could thus potentially provide essential insights into the circuit mechanism of basal ganglia underlying action selection.

Linear and nonlinear control of action selection by direct vs. indirect pathway

To gain an overall picture of how basal ganglia control action selection, we run through the model with a wide continuous range of manipulation to mimic the effects from lesion to optogenetic inhibition and optogenetic activation (Figure 7A, B, E, F). The simulations of cell ablation and bidirectional optogenetic manipulations of D1-SPNs activity in the model reveal no significant effects at 2-s trials (Figure 7C), but a linear relationship between the neuronal activity in direct pathway and the behavioral performance of choice in 8 s trials (Figure 7D), as observed in animal experiments. It thus further confirms that direct pathway selects action and facilitates ongoing choice, consistent with the predictions from both the classic Go/No-go and recent Co-activation models (Figure 7I, Figure S9A-D) (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996).

Computational modeling reveals direct and indirect pathways regulating action selection in a distinct manner.

(A) Schematic for manipulation of D1-SPNs in ‘Triple-control’ model. (B) Schematic of manipulation of D1-SPNs in the center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (C) Correct rate change in 2s trials when manipulating D1-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 1.171, p = 0.238). (D) Correct rate change in 8s trials when manipulating D1-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 13.71, p < 0.0001). (E) Schematic for optogenetic manipulation of D2-SPNs in ‘Triple-control’ model. (F) Schematic of manipulation of D2-SPNs in the center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (G) Correct rate change in 2s trials when manipulating D2-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 59.13, p < 0.0001). (H) Correct rate change in 8s trials when manipulating D2-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 40.75, p < 0.0001). (I) Diagram of linear modulation of direct pathway. (J) Diagram of nonlinear modulation of indirect pathway.

In contrast, manipulations of D2-SPNs activity from cell ablation to optogenetic inhibition and then optogenetic stimulation in the model demonstrate an inverted-U-shaped nonlinear relationship between the neuronal activity in indirect pathway and action selection, for both 2-s and 8-s trials (Figure 7G, H). Detailed analyses reveal that D2-SPNs ablation removes both the Co-activation and Go/No-go module in the indirect pathway and leaves SNr activity dictated by D1-SPN inputs. However, due to the inhibition from Co-activation to Go/No-go module in the indirect pathway via D2-SPN collaterals and short-term plasticity of these synapses (Figure S7A-C; see Methods), optogenetic manipulation of D2-SPNs differentially affects the D2-SPN subpopulations groups and promotes Go/No-go module to dominate the basal ganglia network (Figure S7C-S). This dynamic switch of dominance between Co-activation and Go/No-go modules on the basal ganglia network gives rise to a nonlinear relationship between D2-SPNs manipulation and the behavioral outcome (Figure 7J).

Note that when the same inputs were applied to the Go/No-go or Co-activation model alone, the behavioral performance in either model exhibits linear negative (Figure S9E) or positive correlation (Figure S9F) with D2-SPNs activity, respectively. Both our experimental and modeling results thus indicate that different from either the Go/No-go or Co-activation model, the indirect pathway regulates action selection in a nonlinear manner, depending on the state of the network and D2-SPNs activity level. Besides collaterals within D2-SPNs, other collateral connections, for example connections between D1-SPNs or connections between D1- and D2-SPNs, could also contribute to the regulation of action selection (Taverna et al., 2008). We tested our ‘Triple-control’ model with adding additional collateral connections as D1→D1 (Figure S10A-C), D1→D2 (Figure S10D-F) and D2→D1 (Figure S10G-I), respectively. It was found while these additional collaterals further quantitatively regulate action selection, the general principle of linear vs. nonlinear modulation of action selection by direct and indirect pathways still qualitatively hold (Figure S10). Interestingly, our current ‘Triple-control’ model can also replicate the behavioral effects of optogenetic manipulation of nigrostriatal dopamine on behavioral choice (Howard et al., 2017), and further unveils an inverted-U-shaped relationship between striatal dopamine concentration change and action selection (Figure S11). Together, these results suggest that there are multiple levels of interactions from D1- and D2-SPNs to dynamically control SNr output, and the basal ganglia direct and indirect pathways distinctly control action selection in a linear and nonlinear manner, respectively.

Discussion

Here, by using an internally-driven 2-8s action selection task in mice, we investigated the function of basal ganglia direct and indirect pathways in mediating dynamic action selection. We found that the neuronal activities in SNr, the major output of basal ganglia, directly reflect animals’ internal action selection process, other than simply time or value. It was also observed that the striatum, the main input of basal ganglia, shares the similar action selection-related neuronal dynamics with SNr and is needed for both learning and execution of proper action selection. Furthermore, the striatal direct and indirect pathways exhibit distinct neuronal activity and during manipulation, they have different functional effects on controlling action selection. Notably, the experimental observations on the physiology and function of direct and indirect pathways cannot be simply explained by either the traditional ‘Go/No-go’ model or the more recent ‘Co-activation’ model. We proposed a new ‘Triple-control’ functional model of basal ganglia, suggesting a critical role of dynamic interactions between different neuronal subpopulations within the indirect pathway for controlling basal ganglia output and behavior. In the model, a ‘center (direct pathway) – surround (indirect pathway) – context (indirect pathway)’ three layers of structure exerts dynamic control of action selection, depending on the input level and network state. This new model respects the realistic neuroanatomy, and can recapitulate and explain the essential in vivo electrophysiological and behavioral findings. It also provides a new perspective on understanding many behavioral phenomena involving in dopamine and basal ganglia circuitry in health and disease.

Our current 2-8s action selection task offers a unique opportunity to observe the animal’s internal switch from one choice to another and monitor the underlying neuronal dynamics correspondingly. We observed two major types of monotonically-changing SNr neuronal dynamics during the internal choice switching, presumably one type associated with selecting one action and another with selecting the competing action, respectively. The classic view on SNr activity is that it tonically inhibits the downstream motor nuclei and releases action via disinhibition (Albin et al., 1989; DeLong, 1990; Hikosaka and Wurtz, 1983; Wurtz and Hikosaka, 1986). The increased response in SNr, however, could potentially inhibit the competing actions or the movements toward the opposing direction through projections to the contralateral brain regions like superior colliculus (Jiang et al., 2003). Here, we found that two subpopulations in SNr showed opposite monotonic firing change during the left-then-right choice, and notably, their neuronal dynamics switched when the animals performed the reversed version of task which requires a right-then-left choice. It thus suggests that these SNr neurons are indeed associated with different action options during choice behavior, and actively adjust their firing rates to facilitate respective action selection. Given the opposite neuronal dynamics and functionally antagonistic nature of these two groups of SNr neurons, we defined the net output of basal ganglia by the subtracting the neuronal activity between the two SNr subpopulations and correlated it with the behavioral choice. We found that the basal ganglia net output exhibited a tight correlation with the psychometric curve of behavioral choice, and faithfully represented a neural basis for the dynamic action selection process.

As one of the major input nuclei of basal ganglia, striatum influences SNr activity through direct/indirect pathways and undisputedly, plays an essential role in action selection (Ding and Gold, 2012; Geddes et al., 2018; Jin et al., 2014; Lauwereyns et al., 2002; Tai et al., 2012). By genetic manipulation and pharmacological inactivation, we showed that striatum is indispensable for both learning action selection and the proper performance of learned behavioral choice. The recording of striatal neuronal activity during action selection further revealed that striatal spiny projection neurons share the similar types of neuronal dynamics as SNr. Through optogenetic-tagging in freely behaving mice, we further found that striatal SPNs in the direct and indirect pathways show distinct activity profile, with D1-SPNs representing a strong bias towards the preferred choice, while D2-SPNs encoding two choices equally.

Two prevailing models have been proposed to explain the functional distinction between D1- and D2-SPNs. The canonical model of the basal ganglia suggests that the D1- and D2-SPNs play antagonistic roles in controlling action as mediating “Go” and “No-go” signals, respectively (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). A more recent model, however, implies that as D1-SPNs initiate an action, D2-SPNs co-activate with D1-SPNs to inhibit other competing actions (Cui et al., 2013; Hikosaka et al., 2000; Isomura et al., 2013; Jin et al., 2014; Mink, 1996). Essentially, these two models agree upon the functional role of D1-SPNs in releasing or facilitating the desired action, but contradict on the function of D2-SPNs on which targeted action of inhibiting. Here, our in vivo recording data indicate that both D1- and D2-SPNs share similar neuronal dynamics during action selection, and the neural activity alone is not sufficient to separate and determine whether “Go/No-go” or “Co-activation” model is supported (Cui et al., 2013; Isomura et al., 2013; Jin et al., 2014). To resolve the functional distinction of the direct vs. indirect pathway, we applied a series of cell-type-specific manipulations on striatal D1- and D2-SPNs during action selection behavior. First, we generated mutant mice in which NMDA receptors are deleted from either striatal D1- or D2-SPNs (Geddes et al., 2018; Jin et al., 2014). Both the D1-NR1 KO and D2-NR1 KO mice showed learning deficits and behavioral choice impairments when tested with probe trials, suggesting that both D1- and D2-SPNs are necessary for learning and performing action selection. Notably, while the D1-NR1 KO mice are mostly impaired in the choice associated with 8s, a less-preferred option compared to 2s, the D2-NR1 KO mice are compromised in both 2s and 8s choice. Additional experiments with cell-type specific ablation further confirmed these results, consistent the distinct neuronal activity profile in these two pathways revealed during in vivo neuronal recording. While both the “Go/No-go” and “Co-activation” models predict the suppression of D1-SPNs activity leads to impaired action selection, supported by current KO and cell-ablation data, the manipulation experiments on D2-SPNs favor the “Co-activation” but not the “Go/No-go” model which the latter suggests D2-SPNs ablation would improve rather than impair action selection.

Next, we directly introduced transient bidirectional manipulations to D1- and D2-SPNs activity by optogenetics while mice performing the task. Our findings revealed that activation or inhibition of D1-SPNs increased and decreased the correct rate of choice respectively, suggesting a facilitating role of direct pathway in action selection, which again fits well with the “Go/No-go” as well as the “Co-activation” model. In contrast, optogenetic activation of D2-SPNs decreased the correct rate of choice, while inhibition of D2-SPNs promoted the correct choice. These results were supportive to the “Go/No-go” model but contradicted to the prediction of “Co-activation” theory, which the latter predicts that activation of D2-SPNs inhibits competing actions to facilitate desired choice, whereas inhibition of D2-SPNs releases competing actions and compromises the ongoing choice.

In summary, neither “Go/No-go” nor “Co-activation” models could fully explain the experimental results we found, particularly for experiments on D2-SPNs in the indirect pathway. Through computer simulation, we further demonstrated that a simple additive combination of “Go/No-go” nor “Co-activation” models by linear addition cannot reproduce all the experimental observations. To resolve these theoretical difficulties, we proposed a new center-surround-context “Triple-control” model of basal ganglia pathways for action selection. Specifically, two subpopulations of D2-SPNs in the indirect pathway function as “Co-activation” and “Go/No-go” modules respectively, and an activity-dependent inhibition from “Co-activation” to “Go/No-go” module mediates the dynamic switch between the dominant module depending on the inputs and network state. Due to the dominant “Co-activation” module in the default state, excessive inhibition of D2-SPNs or ablating the entire indirect pathway eliminates the promotive contribution and impairs action selection in the “Triple-control” model, consistent with the experimental observations. In contrast, transient increase of D2-SPNs firing activity during optogenetic stimulation introduces shift toward the “Go/No-go” dominance from the “Co-activation” module via firing-rate-dependent short-term depression of the inhibitory synapses between them, which amplifies the “No-go” signal and compromises action selection as experimentally found. In contrast, transient decrease of D2-SPNs firing activity during optogenetic inhibition results in disinhibition of “Go/No-go” module from inhibitory control of “Co-activation” module, with an attenuated “No-go” signal which leads to better performance in choice. These results from our new “Triple-control” model thus suggest that the basal ganglia circuitry could be much more dynamic than previously thought, and it could employ a complex mechanism of functional module reconfiguration for context- or state-dependent flexible control.

In the “Triple-control” model, we posited the collateral connections among striatal D2-SPNs and its short-term plasticity could serve as an operational mechanism for the dominant module switching. However, besides these well-known striatal local connections as one of the simplest possible mechanisms, other anatomical circuits within basal ganglia circuitry could potentially fulfill this functional role alone or additionally as well. For example, striatal D2-SPNs project to external global pallidus (GPe) through striatopallidal pathway, and meanwhile they receive arkypallidal projections from GPe to both the striatal SPNs and interneurons (Abdi et al., 2015; Fujiyama et al., 2016; Mallet et al., 2016). It is thus also possible that the dynamic interaction between “Co-activation” and “Go/No-go” modules is mediated through di-synaptic or tri-synaptic modulation with GPe and/or striatal interneurons involved. Furthermore, in theory this dynamic interaction between “Co-activation” and “Go/No-go” modules can also occur outside striatum in the downstream nuclei including GPe and SNr, given their specific neuronal subpopulations receiving inputs from corresponding striatal D2-SPNs subgroups and proper collateral connections within the nuclei (Atherton et al., 2013; Cazorla et al., 2014; Fujiyama et al., 2011; Lee et al., 2020; Wu et al., 2000). Considering the crucial role of dopamine in basal ganglia circuitry, the new “Triple-control” model can also reproduce our previous experiments results on the effect of nigrostriatal dopamine on action selection (Howard et al., 2017). Importantly, it unveils that there is an inverted-U-shaped relationship between dopamine concentration change and action selection (Cools and D’Esposito, 2011). The model simulation suggests while moderate dopamine increase improves decision making, too much dopamine changes, either increase or decrease, dramatically impairs the choice behavior. These results might be able to explain some of behavioral observations involving in obscure decision making under the influence of addictive substances.

Our findings also have important implications in many neurological and psychiatric diseases. It was known that the loss of dopamine leads to hyperactivity of D2-SPNs and disruption of local D2-SPNs collaterals in Parkinson’s disease (Taverna et al., 2008; Wei and Wang, 2016). These alterations will not only break the balance of direct vs. indirect pathway, but also disrupt the multiple dynamic controls from the indirect pathway. The action selection will thus be largely problematic, even with L-DOPA treatment, which might restore the dopamine partially but not necessarily the altered basal ganglia circuitry and its circuit dynamics (Bastide et al., 2015). The current “Triple-control” model also provides some mechanistic insights into the inhibitory control deficits observed Schizophrenia (Taverna et al., 2008; Wei and Wang, 2016). For instance, an increase in the density and occupancy of the striatal D2 receptors (D2R) has been frequently reported in schizophrenia patients (Abi-Dargham et al., 2000; Howes and Kapur, 2009; Laruelle et al., 1997; Wong et al., 1986). Many antipsychotic medications primarily aim to block the D2R (la Fougere et al., 2005; Lally and MacCabe, 2015; Yokoi et al., 2002), but the drug dose is the key to the treatment and severer adverse effects are associated with overdose of D2R antagonism (Levine and Ruha, 2012). In addition, prolonged exposure to antipsychotics often causes extrapyramidal symptoms, including Parkinsonian symptoms and tardive dyskinesia (Jarskog et al., 2007; Seeman, 2002). The dose-dependent effects when modulating D2R were also found in cognitive functions such as serial discrimination, in which relatively low and high dose of D2R agonist in striatum impair the performance in the discrimination task, while the intermediate dose of D2R agonist produces significant improvement (Cools and D’Esposito, 2011; Goldman-Rakic et al., 2000; Horst et al., 2019; Mattay et al., 2003). These observations thus further underscore the dynamic interplays and complexity of basal ganglia pathways in action control, as demonstrated in current study and the new triple-control functional model.

eLife Assessment

This valuable work proposes a new model of basal ganglia function termed the "Triple Control Model" consisting of one direct pathway and two indirect pathways. The study provides evidence that the firing dynamics of substantia nigra reflect action selection but not interval time or reward and makes a case for the need for the Triple Control Model to explain the results of their manipulations of D1 and D2 receptor-expressing cells in the striatum. Although this work may be of interest to neuroscientists who study basal ganglia, without additional analyses of the relationship of neural activity to movements, and direct experimental evidence of two indirect pathways, the evidence for this model is somewhat incomplete.

Reviewer #1 (Public Review):

The manuscript describes an interesting experiment in which an animal had to judge a duration of an interval and press one of two levers depending on the duration. The Authors recorded activity of neurons in key areas of the basal ganglia (SNr and striatum), and noticed that they can be divided into 4 types.

The data presented in the manuscript is very rich and interesting, however, I am not convinced by the interpretation of these data proposed in the paper. The Authors focus on neurons of types 1 & 2 and propose that their difference encodes the choice the animal makes. However, I would like to offer an alternative interpretation of the data. Looking at the description of task and animal movements seen in Figure 1, it seems to me that there are 4 main "actions" the animals may do in the task: press right lever, press left lever, move left, and move right. It seems to me that the 4 neurons authors observed may correspond to these actions, i.e. Figure 1 shows that Type 1 neurons decrease when right level becomes more likely to be correct, so their decrease may correspond to preparation of pressing right lever - they may be releasing this action from inhibition (analogously Type 2 neurons may be related to pressing left lever). Furthermore, comparing animal movements and timing of activity of neurons of type 3 and 4, it seems to me that type 3 neurons decrease when the animal moves left, while type 4 when the animal moves right.

I suggest Authors analyse if this interpretation is valid, and if so, revise the interpretation in the paper and the model accordingly.

Reviewer #2 (Public Review):

In this valuable manuscript Li & Jin record from the substantial nigra and dorsal striatum to identify subpopulations of neurons with activity that reflects different dynamics during action selection, and then use optogenetics in transgenic mice to selectively inhibit or excite D1- and D2- expressing spiny projection neurons in the striatum, demonstrating a causal role for each in action selection in an opposing manner. They argue that their findings cannot be explained by current models and propose a new 'triple control' model instead, with one direct and two indirect pathways. These findings will be of broad interest to neuroscientists, but lacks some direct evidence for the proposal of the new model.

Overall there are many strengths to this manuscript including the fact that the empirical data in this manuscript is thorough and the experiments are well-designed. The model is well thought through, but I do have some remaining questions and issues with it.

Weaknesses:

1. The nature of 'action selection' as described in this manuscript is a bit ambiguous and implies a level of cognition or choice which I'm not sure is there. It's not integral to the understanding of the paper really, but I would have liked to know whether the actions are under goal-directed/habitual or even Pavlovian control. This is not really possible to differentiate with this task as there are a number of Pavlovian cues (e.g. lever retraction interval, house light offset) that could be used to guide behavior.

2. In a similar manner, the part of the striatum that is being targeted (e.g. Figures 4E,I, and N) is dorsal, but is central with regards to the mediolateral extent. We know that the function of different striatal compartments is highly heterogeneous with regards to action selection (e.g. PMID: 16045504, 16153716, 11312310) so it would have been nice to have some data showing how specific these findings are to this particular part of dorsal striatum.

3. I'm not sure how I feel about the diagrams in Figure 4S. In particular, the co-activation model is shown with D2-SPNs represented as a + sign (which is described as "having a facilitatory effect to selection" in the caption), but the co-activation model still suggests that D2-SPNs are largely inhibitory - just of competing actions rather than directly inhibiting actions. Moreover, I am not sure about these diagrams because they appear to show that D2-SPNs far outnumbers D1-SPNs and we know that this isn't the case. I realize the diagrams are not proportionate, but it still looks a bit misrepresented to me.

4. There are a number of grammatical and syntax errors that made the manuscript difficult to understand in places.

5. I wondered if the authors had read PMID: 32001651 and 33215609 which propose a quite different interpretation of direct/indirect pathway neurons in striatum in action selection. I wonder if the authors considered how their findings might fit within this framework.

6. There is no direct evidence of two indirect pathways, although perhaps this is beyond the scope of the current manuscript and is a prediction for future studies to test.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

The manuscript describes an interesting experiment in which an animal had to judge a duration of an interval and press one of two levers depending on the duration. The Authors recorded activity of neurons in key areas of the basal ganglia (SNr and striatum), and noticed that they can be divided into 4 types.

The data presented in the manuscript is very rich and interesting, however, I am not convinced by the interpretation of these data proposed in the paper. The Authors focus on neurons of types 1 & 2 and propose that their difference encodes the choice the animal makes. However, I would like to offer an alternative interpretation of the data. Looking at the description of task and animal movements seen in Figure 1, it seems to me that there are 4 main "actions" the animals may do in the task: press right lever, press left lever, move left, and move right. It seems to me that the 4 neurons authors observed may correspond to these actions, i.e. Figure 1 shows that Type 1 neurons decrease when right level becomes more likely to be correct, so their decrease may correspond to preparation of pressing right lever - they may be releasing this action from inhibition (analogously Type 2 neurons may be related to pressing left lever). Furthermore, comparing animal movements and timing of activity of neurons of type 3 and 4, it seems to me that type 3 neurons decrease when the animal moves left, while type 4 when the animal moves right.

I suggest Authors analyse if this interpretation is valid, and if so, revise the interpretation in the paper and the model accordingly.

We thank the reviewer for the general appreciation of the study. Regarding to the interpretation of each SNr subtypes, we have compared firing activities of the same SNr neurons in both standard 2-8 s task and reversed 2-8 s task (Figure 2G-R, Figure S4). Type 1 and Type 2 neurons are related to right and left choices respectively in the standard task (Figure 2G, M, N), and this is even more evident in the reversed 2-8 s task (Figure 2J), because when the movement trajectories of the same mice in 8-s trials were reversed from left-then-right in the control task (Figure 2I) to right-then-left in the reversed task (Figure 2L), the Type 1 SNr neurons which showed monotonic decreasing dynamics in the control 2-8 s task (Figure 2M) reversed their neuronal dynamics to a monotonic increase in the reversed 2-8 s task (Figure 2P). The same reversal of neuronal dynamics was also observed in Type 2 SNr neurons in the reversed version of standard task (Figure 2N vs Figure 2Q). Therefore, Type 1 and Type 2 neurons are related to the action selection. Furthermore, Type 3 and Type 4 SNr neurons exhibiting transient change when mice switching either from left to right, or from left to right maintained the same neuronal dynamics in both standard 2-8 s task and reversed 2-8 s task (Figure S4C-F), indicating that Type 3 and Type 4 neurons are related to the switch between choices but not the specific upcoming choice to be made.

Reviewer #1 (Recommendations For The Authors):

Suggest to clarify if SNr neurons recorded just from a single hemisphere or bilaterally.

We have described the recording hemisphere in our Methods (page 46, lines 974-976) as follows “For striatum recording, we implanted 11 mice in the left hemisphere and 8 mice in the right hemisphere. For the SNr recording, we implanted 5 mice in the left hemisphere and 4 mice in the right hemisphere.”

Suggest to analyse if type 1/2/3/4 neurons are preferrably located in hemispheres contra/ipsi lateral to a particular lever or movement.

We have addressed this issue in Figure S3 and Figure S6. In fact, we have implanted electrodes in both left and right hemispheres with mirror M-L coordinates. For striatum recording, we implanted 11 mice in the left hemisphere and 8 mice in the right hemisphere. For the SNr recording, we implanted 5 mice in the left hemisphere and 4 mice in the right hemisphere. We have analyzed the striatal and SNr neuronal activity in left vs. right hemisphere respectively, in relation to action selection. We found that SNr neurons recorded in either left or right hemisphere exhibited the same four types of neural dynamics with similar proportions (Fig. S3). Specially, the Type 1 neurons are dominant in both hemispheres. Similar in striatum, SPNs from left and right hemispheres showed the same four types of neural dynamics with similar proportions (Fig. S6). Therefore, there is no significant difference between hemispheres regarding to the proportion of neuron subtypes.

Suggest to investigate if type 1/2 neurons are involved in preparation for lever press, please investigate if these neurons are also changing their activity during the lever press.

In Figure S1L, we have showed the neuronal activities of example Type 1 and Type 2 SNr neurons to rewarded and non-rewarded lever presses. Type 1 SNr neuron shows higher firing activities when pressing the left lever than pressing the right lever, whereas Type 2 SNr neuron shows higher firing activities when pressing the right lever than pressing the left lever, indicating that Type 1 and Type 2 neurons firing activities are action choice dependent.

Suggest investigating if Type 3/4 neurons are controlling movement from one location to another, please analyse if their activity is correlated with the movement on trial by trial bases.

In Figure S2C-D, we showed firing activities of example Type 3 and Type 4 neurons on trial-by-trial bases. Type 3 neuron showed increased firing activities between 3-4 s during the 8s lever retraction period when the animal switched from left side to right side, whereas Type 4 neuron showed decreased firing activities between 3-4 s during as the animal switching from left to right. We further showed in Figure S4C-F, Type 3 and Type 4 neurons Type 3 and Type 4 neurons are related to the switch between choices but not the specific upcoming choice to be made.

Suggest also performing analogous analyses for striatal neurons.

We showed 4 types of SPNs on the on trial-by-trial bases as follows. Due to the limitation of the number of figures, these data were not included in the manuscript. We have now included these results in Fig. S2(E-H).

Typo: l. 68: "can bidirectionally regulates" -> "can bidirectionally regulate"

Thanks, we have now corrected the typos.

Reviewer #2 (Public Review):

In this valuable manuscript Li & Jin record from the substantial nigra and dorsal striatum to identify subpopulations of neurons with activity that reflects different dynamics during action selection, and then use optogenetics in transgenic mice to selectively inhibit or excite D1- and D2- expressing spiny projection neurons in the striatum, demonstrating a causal role for each in action selection in an opposing manner. They argue that their findings cannot be explained by current models and propose a new 'triple control' model instead, with one direct and two indirect pathways. These findings will be of broad interest to neuroscientists, but lacks some direct evidence for the proposal of the new model.

Overall there are many strengths to this manuscript including the fact that the empirical data in this manuscript is thorough and the experiments are well-designed. The model is well thought through, but I do have some remaining questions and issues with it.

Weaknesses:

1. The nature of 'action selection' as described in this manuscript is a bit ambiguous and implies a level of cognition or choice which I'm not sure is there. It's not integral to the understanding of the paper really, but I would have liked to know whether the actions are under goal-directed/habitual or even Pavlovian control. This is not really possible to differentiate with this task as there are a number of Pavlovian cues (e.g. lever retraction interval, house light offset) that could be used to guide behavior.

Sorry for the confusion of task description in the manuscript. We appreciate reviewer’s deep understanding about the complexity of the 2-8 s task we designed. Indeed, the 2-8 s task can’t be simply categorized as goal-directed/habitual or Pavlovian task. There are several behavioral aspects in this task. Lever retraction is served as a Pavlovian cue for mice to start performing the left-then-right sequential movement, but once levers are retracted, there is no cue available to mice during the lever retraction period, and mice have to make a decision to switch choice solely based on its internal estimation of the passage of time, which is considered as a cognitive process. The house light stays on for the entire training session (2 – 3 hours), and will be turned off when the task is done, so house light will not be used as a guidance for choice behavior. The behavior and neural activities during the lever retraction period is our main focus in this manuscript. The main advantage of such task design is that the animal is engaged in a self-determined, dynamic switch of action selection process, which offers a unique opportunity for investigating the role of various neuronal populations in the basal ganglia pathways during action selection.

1. In a similar manner, the part of the striatum that is being targeted (e.g. Figures 4E,I, and N) is dorsal, but is central with regards to the mediolateral extent. We know that the function of different striatal compartments is highly heterogeneous with regards to action selection (e.g. PMID: 16045504, 16153716, 11312310) so it would have been nice to have some data showing how specific these findings are to this particular part of dorsal striatum.

We thank the reviewer for bringing up this point. We are targeting dorsal-central part of striatum. In Figure S5G-L, we showed the specific location we targeted in striatum. Also as specified in Methods (lines 965-970), the craniotomies for electrode implantation were made at the following coordinates: 0.5 mm rostral to bregma and 1.5 mm laterally, and ~ 2.2 mm from the surface of the brain for dorsal striatum. For the virus injection and optic fiber implantation (lines 997-998), the craniotomies was made bilaterally at 0.5 mm rostral to bregma, 2 mm laterally and ~ 2.2 mm from the surface of the brain.

1. I'm not sure how I feel about the diagrams in Figure 4S. In particular, the co-activation model is shown with D2-SPNs represented as a + sign (which is described as "having a facilitatory effect to selection" in the caption), but the co-activation model still suggests that D2-SPNs are largely inhibitory - just of competing actions rather than directly inhibiting actions. Moreover, I am not sure about these diagrams because they appear to show that D2-SPNs far outnumbers D1-SPNs and we know that this isn't the case. I realize the diagrams are not proportionate, but it still looks a bit misrepresented to me.

We appreciate the reviewer’s comments about the diagram. We borrowed and extended the “center-surround” layout from the receptive field of neurons in the early visual system, as an intuitive analogy in describing the functional interaction among striatal pathways (also see Mink 2003 Archives of Neurology). In the co-activation model, if D2-SPNs inhibit the competing action, then the target action will be more likely to be selected due to the reduced competition, which means D2-SPNs actually facilitate the target action in an indirect way. And this is why we define the effect of D2-SPNs in the co-activation model as facilitatory. The area of each region does not represent the amount of cells but mainly qualitative functional role. To make it clearer, we have now added more explanation in the manuscript (page 17, lines 338-341).

4). There are a number of grammatical and syntax errors that made the manuscript difficult to understand in places.

We have now gone through the text carefully and corrected the typos.

1. I wondered if the authors had read PMID: 32001651 and 33215609 which propose a quite different interpretation of direct/indirect pathway neurons in striatum in action selection. I wonder if the authors considered how their findings might fit within this framework.

We appreciate the reviewer’s comments and suggestion. Miriam Matamales et al. (2020, PMID: 32001651) found that dynamic D2- to D1-SPNs transmodulation across the striatum that is necessary for updating previously learned behavior, which highlights the importance of collateral modulations between D1- and D2-SPNs as an additional layer of behavior control besides the classic direct and indirect pathways. This finding is compatible with our “Triple control” model emphasizing the influence of collateral modulations within striatum on behavior choice. James Peak et al. (2020, PMID: 33215609) demonstrated that D2-SPNs are critical to maintain the flexibility of behavior, which is reflected in our “Triple-control” model that activation of D2-SPNs could trigger the behavioral switch from the current action to another action. Although the two studies mentioned above mainly investigate the roles of striatal D1- and D2-SPNs in action learning and behavioral strategies, their functions in general fit within our new ‘Triple-control’ model of basal ganglia pathways for action selection.

1. There is no direct evidence of two indirect pathways, although perhaps this is beyond the scope of the current manuscript and is a prediction for future studies to test.

As accumulating RNA-seq and physiological data implying the heterogeneity of D2-SPNs, the further investigation of the subtypes of D1- and D2-SPNs and their functionality are likely a direction the field will continue to explore. On the other hand, we have discussed other possible anatomical circuits within basal ganglia circuitry that could fulfill the functional role of a third pathway in our new ‘Triple-control’ model, together with or independent of the second indirect pathway (page 32-33, lines 689-700). We certainly hope that our new model will inspire future work to identify and dissect the additional functional pathways in the basal ganglia circuits for action control.

Reviewer #2 (Recommendations For The Authors):

Suggestions for authors:

1. Consider how specific to the dorso-central striatum these findings are, possibly in the discussion.

We have specified in the Discussion that the study is targeting dorsal-central part of striatum (page 29, lines 609-612).

1. Modify the diagrams in 4S to make them more representative of the model's features.

We have responded this comment above.

1. Consider whether the findings here might fit within the role for direct pathway in excitatory action-outcome learning and the indirect pathway in response flexibility more generally.

The current study is mainly focus on selection and execution of actions. It will definitely be important to continue exploring the functionality of direct vs. indirect pathways in the action learning process.

1. Correct typos and grammatical errors including (but not limited to):

a) Line 62-64 - explain why this is controversial? Is it because we don't know which one applies?

In the “Go/No-go” model, indirect pathway inhibits the desired action and function as gain modulation, while in the “Co-activation” model, indirect pathway inhibits the competing action and in turn facilitates the desired action in an indirect manner, therefore these two existing models disagree with each other on the explanation the function of indirect pathway in its targeting action and the net outcome of behavior.

b) Line 68 - Regulates should be regulate.

This has been corrected in the revised manuscript.

c) Line 86 - should read "there are neuronal populations in either the direct or indirect pathway that are activated..."

This has been corrected in the revised manuscript.

d) Line 146-147 - "these types of neuronal dynamics in Snr only appeared in the correct but not incorrect trials" - It seems the authors are suggesting this only for Types 1 and 2 neurons, but this confused me the first time I read it and I suggest it is made clearer.

Line 146-147 now reads “These four types of neuronal dynamics in SNr only appeared…”

e) Line 346 - significant should be significantly.

This has been corrected in the revised manuscript.

f) Line 360 "contrast" should be "contrasting".

This has been corrected in the revised manuscript.

Version 2

Abstract

Summary

The basal ganglia are known to be essential for action selection. However, the functional role of basal ganglia direct and indirect pathways in action selection remains unresolved. Here by employing cell-type-specific neuronal recording and manipulation in mice trained in a choice task, we demonstrate that multiple dynamic interactions from the direct and indirect pathways control the action selection. While the direct pathway regulates the behavioral choice in a linear manner, the indirect pathway exerts a nonlinear inverted-U-shaped control over action selection, depending on the inputs and the network state. We propose a new center (direct) - surround (indirect) - context (indirect) “Triple control” functional model of basal ganglia, which can replicate the physiological and behavioral experimental observations that cannot be simply explained by either the traditional “Go/No-go” or more recent “Co-activation” model. These findings have important implications on understanding the basal ganglia circuitry and action selection in health and disease.

Introduction

Selecting the proper actions is essential for organism’s survival and reproduction in the ever-changing environment (Gallistel, 1980). Numerous studies have implicated that the basal ganglia, a series of interconnected subcortical nuclei including the striatum and substantia nigra, play a primary role in action selection (Graybiel, 1998; Hikosaka et al., 1998; Jin and Costa, 2015; Mink, 2003; Redgrave et al., 1999). Indeed, a wide range of neurological and psychiatric disorders associated with the dysfunctional basal ganglia circuitry, including Parkinson’s disease (Benecke et al., 1987), Huntington’s disease (Phillips et al., 1995), Obsessive-compulsive disorder (OCD) (Graybiel and Rauch, 2000), are characterized by major deficits in action selection and movement control. Anatomically, commands for motor control are processed by basal ganglia through two major pathways, termed direct and indirect pathway, originating from striatal D1- and D2-expressing spiny projection neurons (D1-/D2-SPNs), respectively (Albin et al., 1989; DeLong, 1990). These two pathways collectively modulate substantia nigra pars reticulata (SNr) activity and the basal ganglia output, thus influence behavioral decisions. There are currently two major types of thinking on how the basal ganglia pathways work. An early classic theory has suggested that the basal ganglia direct and indirect pathways oppose each other to facilitate and inhibit action, respectively (the “Go/No-go” model) (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). In contrast, a recent theory has proposed that direct pathway selects the desired action, while the indirect pathway inhibits other competing actions in order to highlight the targeted choice (the “Co-activation” model) (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996).

The two theories have essentially agreed upon the function of direct pathway being the positive driving force for initiating or facilitating the desired actions. Yet, the ideas about the indirect pathway function are largely controversial as either impeding the desired action in the “Go/No-go” model or inhibiting the competing actions in the “Co-activation” model. While the precise neuroanatomy on how the D2-SPNs control SNr through indirect pathway has yet to be mapped out at single-cell level to differentiate the two hypotheses, either theory has found its supports from behavioral and physiological observations. For instance, it has been found that stimulation of striatal direct and indirect pathways can bidirectionally regulate locomotion (Durieux et al., 2012; Kravitz et al., 2010), consistent with the traditional ‘Go/No-go’ model. On the other hand, in vivo electrophysiological and imaging experiments revealed that the striatal direct and indirect pathways are both activated during action initiation (Barbera et al., 2016; Cui et al., 2013; Geddes et al., 2018; Isomura et al., 2013; Jin et al., 2014; Klaus and Plenz, 2016; Markowitz et al., 2018; Nonomura et al., 2018), as the ‘Co-activation’ model predicted. Furthermore, physiological and optogenetic studies concerning complex behavior such as learned action sequences have further complicated the issue, and unveiled various neuronal subpopulations in both pathways are activated during the initiation, termination and switching of actions (Geddes et al., 2018; Jin and Costa, 2015; Jin et al., 2014; Tecuapetla et al., 2016). So far, how exactly the basal ganglia direct and indirect pathways work together to control action selection has been controversial and inconclusive, and the underlying circuit mechanism remains largely unclear (Calabresi et al., 2014).

Here we trained mice to perform an operant action selection task where they were required to select one out of two actions to achieve reward, based on self-monitored time intervals (Howard et al., 2017). By employing in vivo neuronal recording, we found that the net output of two opponent SNr neuron populations is predictive of the behavioral choices. Through identifying striatal pathway-specific neuronal activity with optogenetic tagging, we found that there are neuronal populations in either the direct or indirect pathway that are activated during selecting one action and suppressed during another. Optogenetic inhibition, as well as selective ablation of direct pathway impairs action selection, and optogenetic excitation of direct pathway enhances current choice, confirming a role of direct pathway in facilitating desired actions. Furthermore, optogenetic inhibition of indirect pathway improves action selection and excitation of indirect pathway impairs behavioral choices, as predicted by the ‘Go/No-go’ model. However, selective ablation of indirect pathway impairs action selection, opposite from the behavioral effect of optogenetic inhibition and at odds with the ‘Go/No-go’ model, but consistent with the prediction from the ‘Co-activation’ model. To resolve these contradictions, we propose a new center (direct) - surround (indirect) - context (indirect) “Triple-control” functional model of basal ganglia pathways, in which there are two interacting indirect pathway subcircuits exerting opposite controls over the basal ganglia output. The new model can reproduce the neuronal and behavioral experimental results that cannot be simply explained by either the “Go/No-go” or the “Co-activation” model. Further systematic analyses from this new model suggested that the direct and indirect pathways modulate behavioral outputs in a linear and nonlinear manner, respectively. Notably, in the new ‘Triple-control’ model, the direct and indirect pathways can work together to dynamically control action selection and operate in a manner similar to ‘Go/No-go’ or ‘Co-activation’ model, depending on the activity level and the network state. These results revise our current understanding on how the basal ganglia control actions, and have important implications for a wide range of movement and psychiatric diseases where the dynamic balance between the two pathways is compromised (Albin et al., 1989; Benecke et al., 1987; Calabresi et al., 2014; DeLong, 1990; Graybiel and Rauch, 2000; Mink, 1996; Phillips et al., 1995).

Results

Opponent SNr activities underlie action selection

To address the role of basal ganglia in action selection, we trained mice in a recently developed 2-8 s task in which they are required to choose the left versus right action based on self-monitored time intervals (Howard et al., 2017). Specifically, mice were put into an operant chamber with both left and right levers extended (Figure 1A, see Methods). For a given trial, both levers retract at trial initiation, and after either 2 s or 8 s (50% for each, randomly interleaved), both levers extend. The mouse has to judge the interval between lever retraction and extension as 2 s vs. 8 s and make a corresponding action choice by pressing the left vs. right lever, respectively (Figure 1A). The first lever press after lever extension was registered as the mouse’s choice. The correct choice leads to sucrose delivery (10 µl) as reward, and any lever presses beyond the first press after lever extension yield no outcome. The animal only has one chance to select the correct choice and gets rewarded in a given trial. If the animal’s very first press after levers extension is the wrong choice, then there’s no reward, and the chance to get rewarded in this particular trial vanishes, or the trial is functionally “terminated” although both levers still available to press. The animal has no second chance to correct its wrong choice by pressing the correct lever after the wrong choice. During the 2s vs. 8s waiting period with lever retraction, the levers are not physically accessible to the animal. Even the animal is trying to approach to the lever during lever retraction, but no lever press will be generated (see Supplemental Video 1). A new trial starts at lever retraction again after a random inter-trial-interval (ITI, 30 s on average; Figure 1A). Across 14 consecutive days of training, mice (n = 10) significantly increased the correct rate of choice from chance level to more than 90% (Figure 1B). In addition, the animals gradually shortened the choice latency and demonstrated a strong preference toward the left lever due to its association with the shorter waiting time (Figure S1A, B). As a result, during the longer-waiting 8 s trials the mouse initially moved toward the left lever, then crossing the midpoint between left and right levers at around 4 s, and stayed around the right lever afterwards (Howard et al., 2017) (Figure 1C; Supplemental Video 1). Note that, the mouse showed no stereotyped movement trajectories during the incorrect trials (Figure 1C). This emerged stereotyped movement trajectory in the 8-s trials thus provided us a unique opportunity for investigating the neural mechanisms underlying the internally-driven, dynamic action selection process.

The neuronal dynamics in SNr during the 2-8 s action selection task.

(A) Schematic diagram for the design of 2-8 s task. (B) Correct rate for wild type mice across 14 days training (n = 10 mice, one-way repeated-measures ANOVA, significant effect of training days, F13,117 = 32.54, p < 0.0001). (C) Movement trajectory of an example mouse in correct (left panel) and incorrect (right panel) 8-s trials (gray line: trajectory of each trials; red/black line: the average trajectory). (D) Diagram of electrode array implanted into substantia nigra pars reticulata (SNr). (E) Firing Rate Index (FRI) of neuronal activity for all task-related SNr neurons in correct 8-s trials. The magnitude of FRI is color coded and the SNr neurons are classified as four different types based on the activity dynamics. (F-I) Averaged FRI for Type 1 (F, green squares indicating activities related to left choice), Type 2 (G, green squares indicating activities related to left choice), Type 3 (H), Type 4 (I) of SNr neurons in correct (red) and incorrect 8-s trials (gray). (J) The proportion of four types of SNr neurons. Type 1 and Type 2 are major types and significantly more than Type 3 and Type 4 (Z-test, p < 0.05). (K) Integrated SNr output defined as the subtraction of averaged FRI between Type 1 and Type 2 SNr neurons. (L) Averaged psychometric curve (n = 10 mice) of choice behavior. (M) The correlation between the Type 1 and Type 2 FRI subtraction and the behavioral choice (R = 0.98, p < 0.0005). Error bars denote s.e.m., same for below unless stated otherwise.

The substantia nigra pars reticulata (SNr) is one of the major output nuclei of basal ganglia (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). To investigate how the basal ganglia contribute to the dynamic process of action selection, we began by recording the SNr neuronal activity in mice trained in the 2-8 s task (Figure 1D, Figure S1C, see Methods). It was found that a large proportion (211/261, 80.8%; recorded from n = 9 mice) of SNr neurons changed firing rate significantly during the correct 8-s trials as mice dynamically shifted the internal action selection from the left to the right (Figure 1E). The Z-score of the task related neuronal firing rate, reflecting the firing activity changes related to baseline, was defined as Firing Rate Index (FRI, see Methods). We focus on the data analyses in the 8-s trials since the first 2-s of 8-s trials consists of the identical behavioral and neuronal profiles of the 2-s trials due to the task design (Figure 1C, Figure S1D-F). The task-related SNr neurons were categorized into four subtypes based on the dynamics of FRI in the correct 8-s trials: Type 1 – monotonic decrease (Figure 1E, F, 102/211, 48.3%), Type 2 - monotonic increase (Figure 1E, G, 56/211, 26.5%), Type 3 - transient phasic increase (Figure 1E, H, 25/211, 11.9%) and Type 4 - transient phasic decrease (Figure 1E, I, 28/211, 13.3%). These four types of neuronal dynamics in SNr only appeared in the correct but not the incorrect trials (Figure 1F-I), nor on the day 1 of task training (Figure S1G-K), suggesting a tight correlation between the SNr neuronal dynamics and the behavioral performance. Here we show trial-by-trial firing activities of SNr example neurons in correct 8s trials from well-trained animals as follows. Although the time of initial approach to the left side varies across trials, trial-by-trial analysis showed that the firing activities are consistent across trials and the averaged activities faithfully reflect the dynamics of each trial, evident for all four types of neurons (Figure S2A-D). Specifically, the Type 1 and Type 2, but not the Type 3 and Type 4 neurons, exhibit firing changes co-varying with the action selection and these two types together consist in around 80% of all task-related SNr neuron population (Figure 1J, Figure S2A-D). There is no dramatic difference in dynamic subtypes and proportion between SNr neurons recorded in left and right hemispheres (Figure S3). Notably, for Type 1 neurons, the firing activities is much higher as animals selected left side at the correct 8s trials than the firing activities when animals selected left side at the incorrect 8s trials (Figure 1F, green squares). The same for Type 2 neurons, their firing activities are dramatically different when animals selected left side in the correct and incorrect trials (Figure 1G, green squares). Therefore, Type 1 and Type 2 dynamics cannot be simply explained by sensory or position related neural activity. Furthermore, we compare the SNr neuron responses at rewarded and non rewarded lever presses. The SNr neuron activities are aligned to lever press at 0 as shown below. For Type 1 SNr neurons, the firing activity at the rewarded left lever presses (defined as the left lever press in correct 2s trials) is much higher than the firing activity at the non-rewarded left lever presses (defined as left lever presses in incorrect 8s trials and random left lever press during the inter-trial-interval). The firing activity difference can also be observed between the rewarded and non-rewarded right lever presses in Type 1 SNr neuron (Figure S1L). For Type 2 SNr neurons, although there’s no difference between the rewarded and non-rewarded left lever presses, the firing activity at the rewarded right lever presses is higher than the firing activity at the non-rewarded right lever presses (Figure S1L). Again, given the same sensory inputs and spatial location for both rewarded and non-rewarded left presses, the difference between rewarded and non-rewarded lever presses indicates that the neural dynamics are action selection dependent, and not simply related to sensory or position information.

It has been suggested that SNr suppresses movements through the inhibition of downstream motor nuclei and releases action via disinhibition (Hikosaka et al., 2000; Mink, 1996). We thus ask whether the opponent neuronal dynamics in Type 1 and Type 2 SNr subpopulations mediate the dynamic shift of choice, by suppressing the competing selection of right vs. left action, respectively. Indeed, the SNr net output by subtracting Type 2 and Type 1 SNr neuronal dynamics (Figure 1K) is highly reminiscent of the animal’s stereotyped movement trajectory during choice (Figure 1C). To further determine the relationship between the SNr net output and action selection, we tested the behavioral choice of the 2-8 s trained mice in a series of non-rewarded probe trials with novel intervals of 2.5, 3.2, 4, 5 and 6.3 s (see Methods). Consistent with what reported before (Howard et al., 2017), the probability of mice selecting the action associated with the long duration (8 s) gradually increases along with the time intervals of probe trials (Figure 1L). The resulting psychometric curve thus represents the animal’s real-time action selection process during the 8-s trials. Further comparison between the psychometric curve and the SNr net output revealed a strong linear correlation (Figure 1M), indicating that the SNr net output faithfully predicts momentary behavioral choice. Together, these results suggest that mice can learn to dynamically shift their choice based on internally-monitored time, and the opponent neuronal activities in SNr correlate with the action selection.

SNr neuronal dynamics reflect action selection but not simply time or value

In the 2-8 s task, the passage of time and expectation of reward both change simultaneously with the animal’s internal choice. One may argue that the Type 1 and Type 2 neuronal dynamics observed in SNr during the 8-s trials might reflect the passage of time or value of expected reward rather than action selection. To differentiate these possibilities and specify the functional role of SNr activity, we presented mice previously trained in the 2-8 s task with random probe trials of 16-s interval (Figure 2A). In these 16-s probe trials which they have never experienced before during training, the animals sometimes wait on the right side and press the right lever, or shift back to the left side and press the left lever when the levers are extended at 16 s (Figure 2B, C). This arbitrary choice situation in the 16-s probe trials thus provides a special window to determine the functional relationship between SNr activity and behavioral choice. If the Type 1 and Type 2 SNr subpopulations encode information about time passage or expectation value, their neuronal activities would continue changing monotonically between 8 and 16 s. In contrast, if the Type 1 and Type 2 SNr subpopulations encode action selection, their neuronal activities would predict the behavioral choice and differentiate between the right vs. left action selection. Indeed, it was found that when the firing activity of Type 1 SNr neurons maintained below baseline from 8 to 16 s, the mice tended to select the right lever later (Figure 2D). However, when the firing activity reversed the decreasing tendency to increase, the mice chose the left lever instead (Figure 2D). A similar relationship between the neuronal activity and behavior choice was also evident in Type 2 SNr neurons, albeit with opposite dynamics (Figure 2E). This is especially evident in the subtraction of Type 2 and Type 1 SNr neuronal dynamics, in which the SNr net output is strongly correlated with and predictive of behavioral choice (Figure 2F). These results thus suggested that the neuronal activities in SNr likely encode the ongoing action selection but not simply reflect time passage or reward value.

SNr neuronal dynamics reflect action selection but not interval time or reward value.

(A) Task diagram of 2-8 s control task with 10% 16-s probe trials. (B) Percentage of behavioral choice in 2-s, 8-s and 16-s trials (blue: left choice; red: right choice) (n = 9 mice, paired t-test, p < 0.05). (C) Movement trajectory of an example mouse in 16-s trials (blue: left choice; red: right choice). (D) Averaged SNr Type 1 FRI in 16-s trials (red: left choice; black: right choice). Firing rates from 8s to 16s (highlighted area) are compared between left and right choice (n = 26 neurons, two-way repeated-measures ANOVA, significant difference between left and right choices, F1,25 = 6.646, p = 0.016). (E) Averaged SNr Type 2 FRI in 16-s trials (red: left choice; black: right choice). Firing rates from 8s to 16s are compared between left and right choice (n = 16 neurons, two-way repeated-measures ANOVA, significant difference between left and right choices, F1,15 = 5.785, p = 0.029). (F) Subtraction of FRI for SNr Type 1 and Type 2 neurons in 16-s probe trials (red: left choice; black: right choice). (G) Task design of 2-8 s standard task. (H) Percentage of behavioral choice in 2-s and 8-s trials (blue: left choice; red: right choice) (n = 6 mice, paired t-test, p < 0.05). (I) Movement trajectory of an example mouse in 8-s trials (blue: left choice; red: right choice). (J) Task design of reversed 2-8 s task. (K) Percentage of behavioral choice in 2-s and 8-s trials in the reversed 2-8 s task (blue: left choice; red: right choice) (n = 6 mice, paired t-test, p < 0.05). (L) Movement trajectory of the same mouse as (I) in 8-s trials in the reversed 2-8 s task (blue: left choice; red: right choice). (M) Averaged FRI of the SNr Type 1 neurons in correct 8-s trials (n = 14 neurons). (N) Averaged FRI of the SNr Type 2 neurons in correct 8-s trials (n = 11 neurons). (O) Integrated SNr output as the subtraction of FRI for SNr Type 1 (M) and Type 2 neurons (N) in the standard 2-8 s task. (P) Averaged FRI of the same neurons as (M) in correct 8-s trials of the reversed 2-8 s task. (Q) Averaged FRI of the same neurons as (N) in correct 8-s trials of the reversed 2-8 s task. (R) Integrated SNr output as the subtraction of FRI for SNr Type 1 (P) and Type 2 neurons (Q) in the reversed 2-8 s task.

To further confirm this point, we recorded the firing activity from the same SNr neurons during both the 2-8 s control task (Figure 2G, 2s-left and 8s-right) and a modified version of 2-8 s task in which the contingency between action and interval is reversed (Figure 2J, 2s-right and 8s-left) on the same day (see Methods). It was found that the mice performed at around 80% correct in both tasks on the same day (Figure 2H, K, Figure S4A). Accordingly, the movement trajectories of the same mice in 8-s trials were reversed from left-then-right in the control task (Figure 2I) to right-then-left in the reversed task (Figure 2L). The left-lever preference during the ITI in the control task was also switched to right-lever preference in the reversed 2-8 s task (Figure S4B). Notably, the passage of time and expected value as well as other environmental factors are all identical in both versions of task, except that the animal’s choice is now reversed from right to left for the 8-s trials (Figure 2H, I vs. Figure 2K, L). If Type 1 or Type 2 SNr neurons encode time or value, either neuronal population will exhibit the same neuronal dynamics in 8-s trials for both versions of task. On the other hand, if Type 1 and Type 2 SNr neurons encode action selection, their neuronal dynamics will reverse in the reversed version of 2-8 s task compared to the standard version. In fact, the Type 1 SNr neurons which showed monotonic decreasing dynamics in the control 2-8 s task (Figure 2M) reversed their neuronal dynamics to a monotonic increase in the reversed 2-8 s task (Figure 2P), consistent with the behavioral choice. The same reversal of neuronal dynamics was also observed in Type 2 SNr neurons in the reversed version of standard task (Figure 2N, Q). The SNr net output by subtracting Type 2 and Type 1 SNr neuronal dynamics, which was tightly correlated with the action selection in the standard 2-8 s task (Figure 2O), is reversed and now predictive of the new behavioral choice in the reversed 2-8 s task (Figure 2R). Notably, Type 3 and Type 4 SNr neurons exhibiting transient change when mice switching between choices maintained the same neuronal dynamics in both tasks (Figure S4C-F). Together, these results therefore demonstrate that the output of basal ganglia reflects the dynamic action selection rather than simply time or value.

Distinct striatal direct vs. indirect pathway activity during action selection

The basal ganglia output is largely controlled by two major neural pathways, called ‘direct’ and ‘indirect’ pathway, originating from D1-vs. D2-expressing spiny projection neurons (D1-vs. D2-SPNs) in the striatum, respectively (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). We then decided to determine the neuronal dynamics in the striatum, specifically the neuronal activity in the direct and indirect pathways during action selection. We employed in vivo extracellular electrophysiology to record the neuronal activity in the dorsal striatum when mice perform the 2-8 s task, and classified putative SPNs based on the spike waveforms and firing properties (Geddes et al., 2018; Jin and Costa, 2010; Jin et al., 2014). Among all the SPNs recorded from the trained mice (n = 19), 341 out of 409 SPNs (83.4%) were defined as task-related neurons for showing significant firing changes during the 2-s and 8-s lever retraction period (Figure 3A, Figure S2E-H, Figure S5A). Similar to the various types of neuronal dynamics observed in SNr, task-related SPNs showed Type 1 (Figure 3B, monotonic decrease, 159/341, 46.6%), Type 2 (Figure 3C, monotonic increase, 103/341, 30.2%), Type 3 (Figure 3D, transient phasic increase, 49/341, 14.4%) and Type 4 (Figure 3E, transient phasic decrease, 30/341, 8.8%) activity profiles during the correct 8-s trials (Figure 3A, Figure S2E-H, Figure S5A). These neural dynamics were largely absent in SPNs on day 1 of training (Figure S5B-F). Also, SPNs recorded from left and right hemispheres showed similar proportions (Figure S6). These results indicate that the striatum, as one of the major input nuclei of basal ganglia, demonstrates the four types of neuronal dynamics similar with SNr during the dynamic process of action selection.

Neuronal activity of striatal D1- and D2-SPNs during action selection.

(A) FRI of neuronal activity for all task-related SPNs in correct 8-s trials. SPNs were classified as Type 1 - 4. (B-E) Averaged FRI for Type 1 (B), Type 2 (C), Type 3 (D), Type 4 (E) of SPNs in correct (red) and incorrect 8-s trials (gray). (F) Diagram of simultaneous neuronal recording and optogenetic identification of D1-vs. D2-SPNs in dorsal striatum. (G) Top panel: Raster plot for a representative D1-SPN response to 100 ms optogenetic stimulation. Each row represents one trial and each black dot represents a spike. Bottom panel: Peristimulus time histogram (PETH) aligned to light onset at time zero. (H) PETH for the same neuron as shown in (G) with a finer time scale. (I) Distribution of light response latencies for D1- and D2-SPNs. (J) Action potential waveforms of the same neuron in (G) for spontaneous (black) and light-evoked (orange) spikes (R = 0.998, P < 0.0001, Pearson’s correlation). (K) Principal component analysis (PCA) of action potential waveforms showing the overlapped clusters of spontaneous (black) and light-evoked (orange) spikes. (L) Proportion of D1-SPN subtypes. Type 1 neurons are significantly more than other three types of neurons in D1-SPNs (Z-test, p < 0.05). (M) Averaged FRI for Type 1 (blue) and Type 2 (red) D1-SPNs in correct 8-s trials. (N) Proportion of D2-SPN subtypes. (O) Averaged FRI for Type 1 (blue) and Type 2 (red) D2-SPNs in correct 8-s trials.

To further determine the neuronal activity in the direct and indirect pathways during action selection, we utilized an optogenetics-aided photo-tagging method (Geddes et al., 2018; Howard et al., 2017; Jin and Costa, 2010; Jin et al., 2014; Lima et al., 2009) to record and identify striatal D1-vs. D2-SPNs in freely behaving mice. Channelrhodopsin-2 (ChR2) was selectively expressed in D1- or D2-SPNs by injecting AAV-FLEX-ChR2 in the dorsal striatum of D1- and A2a-Cre mice, respectively (Geddes et al., 2018; Jin et al., 2014). In the end of each behavioral session with recording, optogenetic stimulation via an optic fiber attached to the electrode array was delivered to identify D1-vs. D2-SPNs through photo-tagging (Figure 3F, Figure S5G-J) (Geddes et al., 2018; Jin et al., 2014). Only those neurons exhibiting a very short latency (≤ 6 ms) to light stimulation (Figure 3G-I) and showing identical spike waveforms (R ≥ 0.95, Pearson correlation coefficient) between behavior and light-evoked response (Figure 3J, K) were identified as Cre-positive thus D1- or D2-SPNs (Geddes et al., 2018; Jin et al., 2014). Within all positively identified D1-SPNs (n = 92 from 6 mice) and D2-SPNs (n = 95 from 6 mice), 74 out of 92 (80.4%) D1-SPNs and 79 out of 95 (83.1%) D2-SPNs showed a significant change in firing rate during the correct 8-s trials. In addition, all four types of neuronal dynamics during action selection were found in both D1-SPNs (Figure 3L, M) and D2-SPNs (Figure 3N, O), as observed in SNr. The Type 1 and Type 2 neuronal dynamics showing monotonic firing change (Figure 3M, O) were the predominant task-related subpopulations within either D1- (Figure 3L) or D2-SPNs (Figure 3N). Notably, the striatal D1-SPNs consist of significantly more Type 1 than Type 2 neurons (Figure 3L), while D2-SPNs show a similar proportion between the two Types (Figure 3N). These data thus suggest while neurons in both the striatal direct and indirect pathways encode information related to behavioral choice, the two pathways might reflect and contribute to distinct aspects of action selection.

Ablation of striatal direct vs. indirect pathway differently impaired action selection

Given the action-selection-related neuronal dynamics observed in striatum, we next asked whether the neural activity in striatum is necessary for learning and execution of action selection, and furthermore, what is the functional difference between the direct and indirect pathways. It has been reported that the NMDA receptors on striatal SPNs are critical for sequence learning (Geddes et al., 2018; Jin and Costa, 2010) and action selection (Howard et al., 2017). To further identify the functional role of NMDA receptors on D1-vs. D2-SPNs for action selection, we employed a genetic strategy to specifically delete NMDA receptors from D1-vs. D2-SPNs by crossing mice carrying a floxed NMDAR1 (NR1) allele with a dorsal striatum-dominant D1-cre line (Gong et al., 2007) and A2a-cre line (Geddes et al., 2018; Jin et al., 2014), respectively (referred to as D1-NR1 KO and D2-NR1 KO mice, respectively; see Methods). Both the D1-NR1 KO and D2-NR1 KO mice are significantly impaired in learning the 2-8 s task compared to their littermate controls (Figure 4A, B), suggesting that NMDA receptors on either D1- or D2-SPNs are critical for learning of proper action selection. In the end of two-week training, when given the probe trials with various intervals across 2 to 8 s, it was found that D1-NR1 KO mice showed a systematic bias toward the lever associated with short interval and made deficient behavioral choice only in long interval trials (Figure 4C). In contrast, D2-NR1 KO mice showed impaired action selection across various probe trials of both short and long intervals (Figure 4D). These data suggest that while NMDA receptors on both D1- and D2-SPNs are required for action learning, the deletion of NMDA receptors in direct and indirect pathways impairs action selection in a different manner.

Selective genetic knockout and ablation of D1- or D2-SPNs distinctly alters action selection.

(A) Correct rate of control (n = 11 mice) and D1-NR1 KO mice (n = 16) in 2-8 s task during 14 days training (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 10.8, p = 0.003). (B) Correct rate of control (n = 17) and D2-NR1 KO mice (n = 10) in 2-8 s task during 14 days training (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 8.728, p = 0.007). (C) The psychometric curve for control (n = 11) and D1-NR1 KO mice (n = 16) (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 12.27, p = 0.002). (D) The psychometric curve for control (n = 17) and D2-NR1 KO mice (n = 10) (two-way repeated-measures ANOVA, significant difference between control and KO mice, F1,25 = 9.64, p = 0.005). (E) Schematic of muscimol infusion into the dorsal striatum in trained mice. (F) Correct rate for control (black: pre-muscimol, gray: post-muscimol) and mice with muscimol infusion (magenta) in dorsal striatum (n = 9 mice, paired t-test, p < 0.01). (G) The psychometric curve for control (n = 9 mice, black: pre-muscimol, gray: post-muscimol control) and mice with muscimol infusion (n = 9 mice, magenta) in dorsal striatum (two-way repeated-measures ANOVA, significant difference between control and muscimol infusion, F2,16 = 11.74, p = 0.0007). (H) Timeline for selective diphtheria toxin (DT) ablation experiments. (I) Schematic of diphtheria toxin receptor (DTR) virus (AAV-FLEX-DTR-GFP) injection in dorsal striatum of D1-Cre mice. (J) Correct rate for control (n = 9 mice) and mice with dorsal striatum D1-SPNs ablation (D1-DTR, n = 8 mice) (two-sample t-test, p = 0.0016). (K) The psychometric curve for control (n = 9 mice) and D1-SPNs ablation mice (n = 8 mice) (two-way repeated-measures ANOVA, main effect of ablation, F1,15 = 1.84, p = 0.195; interaction between trial intervals and ablation, F6,90 = 4.14, p = 0.001). (L) Movement trajectory of a control mouse in 8-s trials. (M) Movement trajectory of a D1-DTR mouse in 8-s trials. (N) Schematic of diphtheria toxin receptor (DTR) virus (AAV-FLEX-DTR-GFP) injection in dorsal striatum of A2a-Cre mice. (O) Correct rate for control (n = 8 mice) and mice with dorsal striatum D2-SPNs ablation (D2-DTR, n = 8 mice) (two-sample t-test, p = 0.005). (P) The psychometric curve for control (n = 9 mice) and D2-SPNs ablation mice (n = 8 mice) (two-way repeated-measures ANOVA, main effect of ablation, F1,15 = 0.477, p = 0.5; interaction between trial intervals and ablation, F6,90 = 12.6, p < 0.001). (Q) Movement trajectory of a control mouse in 8-s trials. (R) Movement trajectory of a D2-DTR mouse in 8-s trials. (S) Schematic of center-surround receptive field diagram for Go/No-Go (left) and Co-activation (right) models. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection.

We then asked whether that neural activity in dorsal striatum is necessary for the proper execution of action selection after learning. We first conduct striatal inactivation in trained wildtype mice by bilateral intra-striatal infusion of muscimol (Figure 4E, see Methods). Striatal muscimol infusion significantly reduced the animal’s overall performance in comparison with the pre- and post-saline injection controls (Figure 4F). When tested with probe trials, the psychometric curve indicated that the inactivation of striatum impairs action selection for the long trials (Figure 4G). These data thus suggested that the striatal neural activity is critical for appropriate execution of learned action selection.

To further elucidate the functional role of specific striatal pathways in action selection, we next employed a viral approach to bilaterally express diphtheria toxin receptors (AAV-FLEX-DTR-eGFP) in the dorsal striatum of trained D1- and A2a-Cre mice, followed by diphtheria toxin (DT) injections to selectively ablate D1- or D2-SPNs (Geddes et al., 2018) (Figure 4H, I, N; see Methods). Ablation of either D1- or D2-SPNs significantly impaired action selection and reduced the correct rate of choice (Figure 4J, O). Notably, the psychometric curve revealed that D1-SPNs ablation mice showed a selective impairment of choice in long interval trials (Figure 4K). In contrast, mice with D2-SPNs ablation exhibited choice deficits in both long and short trials (Figure 4P). Consistent with the D1- and A2a-NR1 KO data, these results suggest that the direct and indirect pathways are both needed yet play distinct roles in action selection.

The classic ‘Go/No-go’ model of basal ganglia suggests the direct and indirect pathways work antagonistically to release and inhibit action, respectively (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). On the other hand, more recent ‘Co-activation’ model of basal ganglia proposes that direct pathway initiates the selected action and at the same time, the indirect pathway inhibits the competing actions (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). For visualization purpose, we diagram ‘Go/No-go’ and ‘Co-activation’ models as center-surround receptive field with D1-SPNs as the center and D2-SPNs as the surround (Figure 4S; Figure S7A, D). The “center-surround” layout is derived from the receptive field of neurons in the early visual system, as an intuitive analogy in describing the functional interaction among striatal pathways (Mink, 2003). The area of each region does not represent the amount of cells but mainly qualitative functional role (Figure 4S). While the direct pathway plays the similar role in both models (Figure S7B, E), the function of indirect pathway differs dramatically (Figure 4S). Lesion of the indirect pathway thus leads to contrast predictions on action selection from the two models (Figure S7C, F). Specifically, ablation of D2-SPNs would facilitate the action being selected through removing inhibition according to the Go/No-go model (Figure S7C) (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010), while blockage of indirect pathway would impair the action selection due to disinhibition of competing actions according to the Co-activation model (Figure S7F) (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). Although our D1-SPNs ablation experiment indicates that direct pathway is required for action selection as suggested in both models (Figure 4J, K), the D2-SPNs ablation result favorably supports the Co-activation model over the Go/No-go model (Figure 4O, P, Figure S7F). In fact, close inspection of the movement trajectories of D1-SPNs lesioned mice in the 8-s trials showed that compared to control mice (Figure 4L), they tend to stick on the left side more often with impaired right choice when lever extension at 8s (Figure 4M). In contrast, D2-SPNs lesioned mice demonstrated overall rather random movement trajectories, and the stereotyped left-then-right movement sequences were largely disrupted in comparison with the controls (Figure 4Q, R). These observations are mostly consistent with the idea of indirect pathway inhibiting competing actions in the Co-activation model (Figure 4S) and lesion of indirect pathway disrupts action selection for both the short and long trials (Figure 4P-R). Together, these data suggest that ablation of direct and indirect pathways both impair choice behavior but in a distinct manner due to their different roles in action selection.

Optogenetic manipulation of D1-vs. D2-SPNs distinctly regulates action selection

To further determine the specific function of direct vs. indirect pathway in action selection, we employed optogenetics to alter the D1- and D2-SPNs activity in vivo with high temporal precision and investigated its effects on the ongoing action selection process. Both the classic ‘Go/No-go’ (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010) and more recent ‘Co-activation’ (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996) models predict that activation of the direct pathway enhances the action selection (Figure S8A, E, I, K, O, Q), while inhibition of direct pathway reduces the correct choice (Figure S8B, F, L, R). To experimentally validate the models’ predictions, AAV-FLEX-ChR2 was injected into the dorsal striatum of D1- or A2a-Cre mice and optic fibers were implanted bilaterally for in vivo optogenetic stimulation (Figure 5A, Figure S5K, L; see Methods) (Geddes et al., 2018; Jin et al., 2014). After mice learned the 2-8s task, 1-s pulse of constant light (wave length 473 nm) was delivered right before lever extension in randomly chosen 50% of 2-s and 50% of 8-s trials (Figure 5B, C, see Methods). The correct rate of optogenetic stimulation trials is compared with the non-stimulation trials of the same animal as a within-subject design. We observed no significant change on the correct rate in 2-s trials, whereas the correct rate was significantly increased by optogenetic stimulation in 8-s trials (Figure 5D), indicating a facilitation effect on action selection by stimulating the D1-SPNs. We then sought to determine the effect of suppressing D1-SPN activity on action selection by viral expression of Halorhodopsin (AAV5-EF1a-DIO-eNpHR3.0-eYFP) in the dorsal striatum of D1-cre mice (Gradinaru et al., 2010). As expected, inhibiting D1-SPNs right before lever extension in trained mice reduced the correct rates in 8-s but not 2-s trials (Figure 5C, E), opposite to D1-SPN stimulation effects. These experimental data with bidirectional optogenetic manipulation suggest that the D1-SPN activity is positively correlated with the choice performance, consistent with the hypothesis of direct pathway facilitating the action selected in both the Go/No-go and Co-activation models (Figure S8K, L, Q, R).

Optogenetic manipulation of D1- vs. D2-SPNs differently regulates action selection.

(A) Schematic of optic fiber implantation for experimentally optogenetic excitation or inhibition of D1- or D2-SPNs in the dorsal striatum. (B, C) Schematic for optogenetic excitation (B) and inhibition (C) of D1-/D2-SPNs for 1 s right before lever extension in 2-8 s task. (D) Change of correct rate for optogenetic excitation of D1-SPNs in 2-s and 8-s trials (n = 11 mice, one-sample t-test, 2-s trials: p = 0.248; 8-s trials: p < 0.05). (E) Change of correct rate for optogenetic inhibition of D1-SPNs in 2-s and 8-s trials (n = 6 mice, one-sample t-test, 2-s trials: p = 0.557; 8-s trials: p < 0.05). (F) Change of correct rate for optogenetic excitation of D2-SPNs in 2-s and 8-s trials (n = 8 mice, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (G) Change of correct rate for optogenetic inhibition of D2-SPNs in 2-s and 8-s trials (n = 5 mice, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (H) Schematic of center-surround receptive field diagram for Go/No-Go (left) and Co-activation (right) models. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection.

Nevertheless, the two models have distinct views on the function of indirect pathway. While the classic ‘Go/No-go’ model suggests that the indirect pathway inhibits the selected action (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010), the ‘Co-activation’ model hypothesizes that the indirect pathway inhibits the competing actions instead (Cui et al., 2013; Hikosaka et al., 2000; Mink, 1996). These models thus provide contrasting predictions about the effect of activation of the indirect pathway on action selection, being decreased correct rate based on the Go/No-go model (Figure S8M) and increased correct rate from the Co-activation model (Figure S8S), respectively. We thus decided to test the distinct predictions from the two models by optogenetic manipulation of indirect pathway during action selection in the 2-8s task. ChR2 or Halorhodopsin (eNpHR3.0) was expressed in the dorsal striatum of A2a-cre mice for bilaterally optogenetic activation or inhibition during behavior (Figure 5A, Figure S5L; see Methods). Notably, optogenetic excitation of D2-SPNs for 1s right before lever extension decreased the correct rate in both 2-s and 8-s trials (Figure 5F). In contrast, transient optogenetic inhibition of D2-SPNs before behavioral choice increased correct rates for both 2-s and 8-s trials (Figure 5G). These data suggest that opposite to the D1-SPN manipulation, optogenetic stimulation of D2-SPNs impairs action selection, while inhibition of D2-SPNs facilitates behavioral choice. These optogenetic results further unveil the distinct roles of direct vs. indirect pathway in action selection, and are in line with the predictions from the Go/No-go (Figure 5H, Figure S8M, N) but not the Co-activation model (Figure S8S, T).

A ‘Triple-control’ model of basal ganglia circuit for action selection

Our DT lesion experiments found that ablation of indirect pathway impairs action selection (Figure 4O, P), as predicted from the Co-activation but not Go/No-go model (Figure 4S), while the optogenetic results suggested that inhibition of D2-SPNs enhances behavioral choice (Figure 5G), a result in favor of the Go/No-go rather than Co-activation model (Figure 5H). We wonder whether these seemly discrepant effects are attributed to a more complex circuit mechanism involving in the indirect pathway different from either the Go/No-go or Co-activation model. To systematically investigate the cell type- and pathway-specific mechanisms underlying action selection, we firstly add Go/No-go and Co-activation models together to examine the whether the resulted combination model could explain the experimental observations (Figure S7G). The lesion of D1-SPNs in the combination model indeed selectively impaired choice in long interval trials (Figure S7H). However, the effect of D2-SPNs ablation in the combination model was neutralized due to the opposing contributions from Go/No-go and Co-activation models respectively (Figure S7I). Based on these simulation results, none of the Go/No-go, Co-activation and combination models was able to fully capture the underlying mechanism of basal ganglia in action selection. Inspired by the data in current experiments, we decided to build a new computational model of the cortico-basal ganglia circuitry based on the realistic neuroanatomy (Aoki et al., 2019; Mailly et al., 2003a; Schmidt and Berke, 2017; Taverna et al., 2008) and empirical neuronal physiology during action selection (Figures 1-3).

Different from the dual control of action by direct vs. indirect pathway in either the Go/No-go or Co-activation model (Figure S7), our new model adds an additional layer of control derived from the indirect pathway, thus called ‘Triple-control’ model for action selection. The combination of Go/No-go or Co-activation models clearly failed to explain all the experimental results (Figure S7G-I), therefore in our model, the new layer of control is not a simple add-on but equipped with interaction with other layers. Specifically, the new model consists of one direct pathway and two indirect pathways defined as D2-SPN #1 and D2-SPN #2 two subpopulations, corresponding to the Co-activation and Go/No-go functional modules, respectively (Figure 6A, B). In addition, the indirect pathway D2-SPNs in the Co-activation module inhibits the indirect pathway D2-SPNs in the Go/No-go module through the well-known D2-SPN collaterals with the properties of short-term depression in the striatum (Gustafson et al., 2006; Schmidt and Berke, 2017; Taverna et al., 2008; Tecuapetla et al., 2007) (Figure 6A; see Methods), providing asymmetric modulation to D2-SPN subgroups and promoting Co-activation module as the dominant functional module at rest. In this ‘Triple-control’ basal ganglia model, striatal D1- and D2-SPNs associated with left and right actions receive excitatory inputs from corresponding cortical inputs (Figure 6A) to generate Type 1 and Type 2 neuronal dynamics (Figure S9A-D) (Lo and Wang, 2006). The D1- and D2-SPNs then regulate the SNr neuronal dynamics through the direct and indirect pathways, respectively (Figure S9) (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996). The net SNr output (Figure S9F, I), which controls the downstream brainstem and thalamic circuits necessary for action selection (Hikosaka, 2007; Lo and Wang, 2006; Redgrave et al., 1999), will determine the final behavioral choice (Figure S9G, J). The choice preference towards left lever over right lever was reflected in the direct pathway by the unevenly weighted connection strength from cortex to D1-SPN Left/Right, as well as the connection strength from D1-SPN Left/Right to SNr Left/Right neurons (see Methods). Our computational simulations showed that this ‘Triple-control’ network model could faithfully recapitulate the neuronal activity across the basal ganglia circuitry and predict the behavioral choice (Figure S9).

A triple-control computational model of basal ganglia direct and indirect pathways for action selection.

(A) Network structure of the cortico-basal ganglia model based on realistic anatomy and synaptic connectivity. (B) Schematic of center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (C) The psychometric curves of behavioral output in control (black) and D1-SPNs ablation condition (blue) in ‘Triple-control’ model (n = 10, two-way repeated-measures ANOVA, main effect of ablation, F1,18 = 98.72, p < 0.0001; interaction between trial intervals and ablation, F6,108 = 7.799, p < 0.0001). (D) The psychometric curves of behavioral output in control (black) and D2-SPNs ablation condition (red) in ‘Triple-control’ model (n = 10, two-way repeated-measures ANOVA, main effect of ablation, F1,18 = 99.54, p < 0.0001; interaction between trial intervals and ablation, F6,108 = 177.6, p < 0.0001). (E) Change of correct rate for optogenetic excitation of D1-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p = 0.407; 8-s trials: p < 0.05). (F) Change of correct rate for optogenetic excitation of D2-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05). (G) Change of correct rate for optogenetic inhibition of D1-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p = 0.28; 8-s trials: p < 0.05). (H) Change of correct rate for optogenetic inhibition of D2-SPNs in 2-s and 8-s trials (n = 10, one-sample t-test, 2-s trials: p < 0.05; 8-s trials: p < 0.05).

To dissect the functional role of direct vs. indirect pathway in action selection, we simulate the cell ablation experiments and examine the behavioral output in the ‘Triple-control’ basal ganglia model. Ablation of D1-SPNs in the network model (Figure S9E) modulates both Type 1 and Type 2 SNr dynamics but in different magnitude due to the biased striatal input to SNr left output and mutual inhibition between SNr left vs. right outputs (Figure S9F; see Methods). As a result, the lesion causes a downward shift in the net SNr output, especially evident at the late section of 8 s (Figure S9G). This change in net SNr output predicts a behavioral bias towards left choice as seen in the psychometric curve (Figure 6C), consistent with experimental results in mice with D1-SPNs ablation (Figure 4K). In contrast, ablation of D2-SPNs in the network model (Figure S9H), by removing the indirect pathways of both the Go/No-go and Co-activation modules, alters Type 1 and Type 2 SNr dynamics (Figure S9I) and change the net SNr output dramatically around 2s as well as 8s (Figure S9J). The model thus predicts behavioral choice deficits for both short and long trials during D2-SPNs ablation (Figure 6D), consistent with experimental observations (Figure 4P). Together, these data suggest that our new ‘Triple-control’ basal ganglia model, based on realistic neuroanatomy and empirical neuronal physiology, can perform action selection similar to the behavior of mice, and successfully replicate the pathway-specific lesion effects on choice.

We further simulate the neuronal and behavioral effects of optogenetic manipulation of D1- and D2-SPNs in the cortico-basal ganglia model. Consistent with the experimental results (Figure 5D, F), optogenetic stimulation of D1-SPNs facilitates the ongoing choice (Figure 6E), while optogenetic inhibition of D1-SPNs suppresses ongoing choice in the model (Figure 6G). In addition, optogenetic stimulation of D2-SPNs impairs the ongoing choice and causes switching (Figure 6F), while optogenetic inhibition of D2-SPNs facilitates ongoing choice, due to the now dominant Go/No-go module mediated by the short-term depression of D2 collaterals in the model (Figure 6H). Consistent with the experimental observations, the optogenetic inhibition effect is opposite from the D2-SPNs cell ablation in the model (Figure 6D).

We next investigate how the striatum influence SNr outputs. Since the collateral projection with STD in D2-SPNs is the key in our ‘Triple-control’ model to switch between Go/No-go and Co-activation modules, we first built a motif of indirect pathway with two D2-SPNs subgroups defined as D2-SPN #1 and D2-SPN #2 (Figure S10A). We tested this indirect pathway motif with monotonic neural dynamics observed in experiments meanwhile simulating the optogenetic activation at 1s and 7s (Figure S10D-I). The SNr therefore received more activation at 1s than at 7s (Figure S10J, K), suggesting that the D2-SPNs with short-term depression in collateral inhibition modulates SNr activities in a firing rate-dependent manner.

We next sought to test the model’s predictions and experimentally investigate the distinctions in modulating SNr activities between the direct and indirect pathways during action selection. In order to manipulate D1- or D2-SPNs and monitor SNr responses at the same time, we simultaneously implanted optogenetic fibers and recording array into striatum and SNr respectively on a single mouse (Figure S10L). While the mice performing the 2-8s task, optogenetic stimulation was delivered to activate either D1- or D2-SPNs. It was found that optogenetic activation of D1- or D2-SPNs caused both inhibition and excitation in Type 1 and Type 2 SNr neurons (Figure S10M-O). To further compare SNr activities responding to striatal activation at different time points during the lever retraction period, for a given trial, we activated D1-SPNs (or D2-SPNs) either at 1s or 7s after the lever retraction (Figure S10P-R). For direct pathway, the change of FRI in SNr activities caused by activation of D1-SPNs showed no significant difference between 1s and 7s (Figure S10P, S). For indirect pathway, activating D2-SPNs at 1s caused smaller activation of FRI than at 7s in Type 1 SNr neurons (Figure S10Q), whereas for Type 2 SNr neurons, activating D2-SPNs at 1s induced bigger FRI increase at 1s than at 7s (Figure S10R). Overall, activating D2-SPNs tended to bias the firing rate downward at 1s but upward at 7s in Type 1 SNr neurons, which was counteractive to the decreasing tendency of Type 1 SNr neuron (Figure S10T). In contrast, Type 2 SNr neurons showed higher FRI increase and smaller decrease in response to activating D2-SPNs at 1s than at 7s, which was opposing to the increasing dynamics of Type 2 SNr neurons (Figure S10T). This firing rate dependent modulation on SNr activities through indirect pathway is consistent with the computational simulation (Figure S10J, K; Figure S11). Therefore, the underlying D2-SPNs collaterals might indeed be a key mechanism contributing to the modulation of SNr activity and action selection in vivo, as simulated in the ‘Triple-control’ model.

Taken together, our new ‘Triple-control’ basal ganglia model, based on realistic neuroanatomy and empirical neurophysiology, successfully reproduces both the lesion and optogenetic data we collected during the animal experiments. It could thus potentially provide essential insights into the circuit mechanism of basal ganglia underlying action selection.

Linear and nonlinear control of action selection by direct vs. indirect pathway

To gain an overall picture of how basal ganglia control action selection, we run through the model with a wide continuous range of manipulation to mimic the effects from lesion to optogenetic inhibition and optogenetic activation (Figure 7A, B, E, F). The simulations of cell ablation and bidirectional optogenetic manipulations of D1-SPNs activity in the model reveal no significant effects at 2-s trials (Figure 7C), but a linear relationship between the neuronal activity in direct pathway and the behavioral performance of choice in 8 s trials (Figure 7D), as observed in animal experiments. It thus further confirms that direct pathway selects action and facilitates ongoing choice, consistent with the predictions from both the classic Go/No-go and recent Co-activation models (Figure 7I, Figure S12A-D) (Albin et al., 1989; DeLong, 1990; Hikosaka et al., 2000; Mink, 1996).

Computational modeling reveals direct and indirect pathways regulating action selection in a distinct manner.

(A) Schematic for manipulation of D1-SPNs in ‘Triple-control’ model. (B) Schematic of manipulation of D1-SPNs in the center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (C) Correct rate change in 2s trials when manipulating D1-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 1.171, p = 0.238). (D) Correct rate change in 8s trials when manipulating D1-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 13.71, p < 0.0001). (E) Schematic for optogenetic manipulation of D2-SPNs in ‘Triple-control’ model. (F) Schematic of manipulation of D2-SPNs in the center-surround-context receptive field diagram for ‘Triple-control’ model. ‘+’ indicates facilitating effect to selection. ‘-’ indicates inhibitory effect to selection. (G) Correct rate change in 2s trials when manipulating D2-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 59.13, p < 0.0001). (H) Correct rate change in 8s trials when manipulating D2-SPNs with different manipulation strengths (n = 10, one-way repeated-measures ANOVA, effect of manipulation strength, F36,324 = 40.75, p < 0.0001). (I) Diagram of linear relationship between the modulation of direct pathway and correct rate of action selection. (J) Diagram of nonlinear relationship between the modulation of indirect pathway and correct rate of action selection.

In contrast, manipulations of D2-SPNs activity from cell ablation to optogenetic inhibition and then optogenetic stimulation in the model demonstrate an inverted-U-shaped nonlinear relationship between the neuronal activity in indirect pathway and action selection, for both 2-s and 8-s trials (Figure 7G, H). Detailed analyses reveal that D2-SPNs ablation removes both the Co-activation and Go/No-go module in the indirect pathway and leaves SNr activity dictated by D1-SPN inputs. However, due to the inhibition from Co-activation to Go/No-go module in the indirect pathway via D2-SPN collaterals and short-term plasticity of these synapses (Figure S10A-C; see Methods), optogenetic manipulation of D2-SPNs differentially affects the D2-SPN subpopulations groups and promotes Go/No-go module to dominate the basal ganglia network (Figure S10C-S). This dynamic switch of dominance between Co-activation and Go/No-go modules on the basal ganglia network gives rise to a nonlinear relationship between D2-SPNs manipulation and the behavioral outcome (Figure 7J).

Note that when the same inputs were applied to the Go/No-go or Co-activation model alone, the behavioral performance in either model exhibits linear negative (Figure S12E) or positive correlation (Figure S12F) with D2-SPNs activity, respectively. Both our experimental and modeling results thus indicate that different from either the Go/No-go or Co-activation model, the indirect pathway regulates action selection in a nonlinear manner, depending on the state of the network and D2-SPNs activity level. Besides collaterals within D2-SPNs, other collateral connections, for example connections between D1-SPNs or connections between D1- and D2-SPNs, could also contribute to the regulation of action selection (Taverna et al., 2008). We tested our ‘Triple-control’ model with adding additional collateral connections as D1→D1 (Figure S13A-C), D1→D2 (Figure S13D-F) and D2→D1 (Figure S13G-I), respectively. It was found while these additional collaterals further quantitatively regulate action selection, the general principle of linear vs. nonlinear modulation of action selection by direct and indirect pathways still qualitatively hold (Figure S13). Interestingly, our current ‘Triple-control’ model can also replicate the behavioral effects of optogenetic manipulation of nigrostriatal dopamine on behavioral choice (Howard et al., 2017), and further unveils an inverted-U-shaped relationship between striatal dopamine concentration change and action selection (Figure S14). Together, these results suggest that there are multiple levels of interactions from D1- and D2-SPNs to dynamically control SNr output, and the basal ganglia direct and indirect pathways distinctly control action selection in a linear and nonlinear manner, respectively.

Discussion

Here, by using an internally-driven 2-8s action selection task in mice, we investigated the function of basal ganglia direct and indirect pathways in mediating dynamic action selection. We found that the neuronal activities in SNr, the major output of basal ganglia, directly reflect animals’ internal action selection process, other than simply time or value. It was also observed that the striatum, the main input of basal ganglia, shares the similar action selection-related neuronal dynamics with SNr and is needed for both learning and execution of proper action selection. Furthermore, the striatal direct and indirect pathways exhibit distinct neuronal activity and during manipulation, they have different functional effects on controlling action selection. Notably, the experimental observations on the physiology and function of direct and indirect pathways cannot be simply explained by either the traditional ‘Go/No-go’ model or the more recent ‘Co-activation’ model. We proposed a new ‘Triple-control’ functional model of basal ganglia, suggesting a critical role of dynamic interactions between different neuronal subpopulations within the indirect pathway for controlling basal ganglia output and behavior. In the model, a ‘center (direct pathway) – surround (indirect pathway) – context (indirect pathway)’ three layers of structure exerts dynamic control of action selection, depending on the input level and network state. This new model respects the realistic neuroanatomy, and can recapitulate and explain the essential in vivo electrophysiological and behavioral findings. It also provides a new perspective on understanding many behavioral phenomena involving in dopamine and basal ganglia circuitry in health and disease.

Our current 2-8s action selection task offers a unique opportunity to observe the animal’s internal switch from one choice to another and monitor the underlying neuronal dynamics correspondingly. We observed two major types of monotonically-changing SNr neuronal dynamics during the internal choice switching, presumably one type associated with selecting one action and another with selecting the competing action, respectively. The classic view on SNr activity is that it tonically inhibits the downstream motor nuclei and releases action via disinhibition (Albin et al., 1989; DeLong, 1990; Hikosaka and Wurtz, 1983; Wurtz and Hikosaka, 1986). The increased response in SNr, however, could potentially inhibit the competing actions or the movements toward the opposing direction through projections to the contralateral brain regions like superior colliculus (Jiang et al., 2003). Here, we found that two subpopulations in SNr showed opposite monotonic firing change during the left-then-right choice, and notably, their neuronal dynamics switched when the animals performed the reversed version of task which requires a right-then-left choice. It thus suggests that these SNr neurons are indeed associated with different action options during choice behavior, and actively adjust their firing rates to facilitate respective action selection. Given the opposite neuronal dynamics and functionally antagonistic nature of Type 1 and Type 2 SNr neurons, we defined the net output of basal ganglia by the subtracting the neuronal activity between the two SNr subpopulations and correlated it with the behavioral choice. The subtraction between Type 1 and Type 2 SNr neurons is the net output of two competing choices and indicates animals’ choice in real time. Also, signals corresponding to left and right choices through direct/indirect pathways eventually converge to SNr (Albin et al., 1989; DeLong, 1990). The collateral inhibition within SNr (Brown et al., 2014; Mailly et al., 2003b) gives rise to the direct competition between different SNr functional subgroups. Therefore, the subtraction between Type 1 and Type 2 SNr neurons represents the outcome of competition between choices. Indeed, we found that the basal ganglia net output exhibited a tight correlation with the psychometric curve of behavioral choice, and faithfully represented a neural basis for the dynamic action selection process.

As one of the major input nuclei of basal ganglia, striatum influences SNr activity through direct/indirect pathways and undisputedly, plays an essential role in action selection (Ding and Gold, 2012; Geddes et al., 2018; Jin et al., 2014; Lauwereyns et al., 2002; Tai et al., 2012). By genetic manipulation and pharmacological inactivation, we showed that striatum is indispensable for both learning action selection and the proper performance of learned behavioral choice. The recording of neuronal activity in dorsal-central striatum during action selection further revealed that striatal spiny projection neurons share the similar types of neuronal dynamics as SNr. Through optogenetic-tagging in freely behaving mice, we further found that dorsal-central striatal SPNs in the direct and indirect pathways show distinct activity profile, with D1-SPNs representing a strong bias towards the preferred choice, while D2-SPNs encoding two choices equally.

Two prevailing models have been proposed to explain the functional distinction between D1- and D2-SPNs. The canonical model of the basal ganglia suggests that the D1- and D2-SPNs play antagonistic roles in controlling action as mediating “Go” and “No-go” signals, respectively (Albin et al., 1989; DeLong, 1990; Kravitz et al., 2010). A more recent model, however, implies that as D1-SPNs initiate an action, D2-SPNs co-activate with D1-SPNs to inhibit other competing actions (Cui et al., 2013; Hikosaka et al., 2000; Isomura et al., 2013; Jin et al., 2014; Mink, 1996). Essentially, these two models agree upon the functional role of D1-SPNs in releasing or facilitating the desired action, but contradict on the function of D2-SPNs on which targeted action of inhibiting. Here, our in vivo recording data indicate that both D1- and D2-SPNs share similar neuronal dynamics during action selection, and the neural activity alone is not sufficient to separate and determine whether “Go/No-go” or “Co-activation” model is supported (Cui et al., 2013; Isomura et al., 2013; Jin et al., 2014). To resolve the functional distinction of the direct vs. indirect pathway, we applied a series of cell-type-specific manipulations on striatal D1- and D2-SPNs during action selection behavior. First, we generated mutant mice in which NMDA receptors are deleted from either striatal D1- or D2-SPNs (Geddes et al., 2018; Jin et al., 2014). Both the D1-NR1 KO and D2-NR1 KO mice showed learning deficits and behavioral choice impairments when tested with probe trials, suggesting that both D1- and D2-SPNs are necessary for learning and performing action selection. Notably, while the D1-NR1 KO mice are mostly impaired in the choice associated with 8s, a less-preferred option compared to 2s, the D2-NR1 KO mice are compromised in both 2s and 8s choice. Additional experiments with cell-type specific ablation further confirmed these results, consistent the distinct neuronal activity profile in these two pathways revealed during in vivo neuronal recording. While both the “Go/No-go” and “Co-activation” models predict the suppression of D1-SPNs activity leads to impaired action selection, supported by current KO and cell-ablation data, the manipulation experiments on D2-SPNs favor the “Co-activation” but not the “Go/No-go” model which the latter suggests D2-SPNs ablation would improve rather than impair action selection.

Next, we directly introduced transient bidirectional manipulations to D1- and D2-SPNs activity by optogenetics while mice performing the task. Our findings revealed that activation or inhibition of D1-SPNs increased and decreased the correct rate of choice respectively, suggesting a facilitating role of direct pathway in action selection, which again fits well with the “Go/No-go” as well as the “Co-activation” model. In contrast, optogenetic activation of D2-SPNs decreased the correct rate of choice, while inhibition of D2-SPNs promoted the correct choice. When stimulating D2-SPNs, animals are still able to press the lever and make a selection shortly after lever extension, therefore, the behavioral effect triggered by D2 stimulation is not simply due to a general effect of decreased locomotion, but the altered action selection process. These results were supportive to the “Go/No-go” model but contradicted to the prediction of “Co-activation” theory, which the latter predicts that activation of D2-SPNs inhibits competing actions to facilitate desired choice, whereas inhibition of D2-SPNs releases competing actions and compromises the ongoing choice.

In summary, neither “Go/No-go” nor “Co-activation” models could fully explain the experimental results we found, particularly for experiments on D2-SPNs in the indirect pathway. Through computer simulation, we further demonstrated that a simple additive combination of “Go/No-go” nor “Co-activation” models by linear addition cannot reproduce all the experimental observations. To resolve these theoretical difficulties, we proposed a new center-surround context “Triple-control” model of basal ganglia pathways for action selection. Specifically, two subpopulations of D2-SPNs in the indirect pathway function as “Co-activation” and “Go/No-go” modules respectively, and an activity-dependent inhibition from “Co-activation” to “Go/No-go” module mediates the dynamic switch between the dominant module depending on the inputs and network state. Due to the dominant “Co-activation” module in the default state, excessive inhibition of D2-SPNs or ablating the entire indirect pathway eliminates the promotive contribution and impairs action selection in the “Triple-control” model, consistent with the experimental observations. In contrast, transient increase of D2-SPNs firing activity during optogenetic stimulation introduces shift toward the “Go/No-go” dominance from the “Co-activation” module via firing-rate-dependent short-term depression of the inhibitory synapses between them, which amplifies the “No-go” signal and compromises action selection as experimentally found. In contrast, transient decrease of D2-SPNs firing activity during optogenetic inhibition results in disinhibition of “Go/No-go” module from inhibitory control of “Co-activation” module, with an attenuated “No-go” signal which leads to better performance in choice. These results from our new “Triple-control” model thus suggest that the basal ganglia circuitry could be much more dynamic than previously thought, and it could employ a complex mechanism of functional module reconfiguration for context- or state-dependent flexible control. More importantly, our model further proposed that while direct pathway regulates action selection in a linear manner, the indirect pathway modulates action selection in a nonlinear inverted-U-shaped way depending on the inputs and the network state (Figure 7). Indeed, the amplitude of activities of D2 pathway is pivotal to the behavioral outcome (Meng et al., 2018). In other words, in certain conditions, activation of D2 pathway will facilitate the action as “start&go”; while too much activation of D2-SPNs will switch D2 pathway to “start&stop” mode (Meng et al., 2018), which is consistent with our proposed “nonlinear” control of D2 pathway over action selection. These results of various functional assemblies defied previous basal ganglia models in which either direct or indirect pathway has been treated as one uniform population and assigned with a single function in controlling action.

In the “Triple-control” model, we posited the collateral connections among striatal D2-SPNs and its short-term plasticity could serve as an operational mechanism for the dominant module switching. However, besides these well-known striatal local connections as one of the simplest possible mechanisms, other anatomical circuits within basal ganglia circuitry could potentially fulfill this functional role alone or additionally as well. For example, striatal D2-SPNs project to external global pallidus (GPe) through striatopallidal pathway, and meanwhile they receive arkypallidal projections from GPe to both the striatal SPNs and interneurons (Abdi et al., 2015; Fujiyama et al., 2016; Mallet et al., 2016). It is thus also possible that the dynamic interaction between “Co-activation” and “Go/No-go” modules is mediated through di-synaptic or tri-synaptic modulation with GPe and/or striatal interneurons involved. Furthermore, in theory this dynamic interaction between “Co-activation” and “Go/No-go” modules can also occur outside striatum in the downstream nuclei including GPe and SNr, given their specific neuronal subpopulations receiving inputs from corresponding striatal D2-SPNs subgroups and proper collateral connections within the nuclei (Atherton et al., 2013; Cazorla et al., 2014; Fujiyama et al., 2011; Lee et al., 2020; Wu et al., 2000). Considering the crucial role of dopamine in basal ganglia circuitry, the new “Triple-control” model can also reproduce our previous experiments results on the effect of nigrostriatal dopamine on action selection (Howard et al., 2017). Importantly, it unveils that there is an inverted-U-shaped relationship between dopamine concentration change and action selection (Cools and D’Esposito, 2011). The model simulation suggests while moderate dopamine increase improves decision making, too much dopamine changes, either increase or decrease, dramatically impairs the choice behavior. These results might be able to explain some of behavioral observations involving in obscure decision making under the influence of addictive substances.

Our findings also have important implications in many neurological and psychiatric diseases. It was known that the loss of dopamine leads to hyperactivity of D2-SPNs and disruption of local D2-SPNs collaterals in Parkinson’s disease (Taverna et al., 2008; Wei and Wang, 2016). These alterations will not only break the balance of direct vs. indirect pathway, but also disrupt the multiple dynamic controls from the indirect pathway. The action selection will thus be largely problematic, even with L-DOPA treatment, which might restore the dopamine partially but not necessarily the altered basal ganglia circuitry and its circuit dynamics (Bastide et al., 2015). The current “Triple-control” model also provides some mechanistic insights into the inhibitory control deficits observed Schizophrenia (Taverna et al., 2008; Wei and Wang, 2016). For instance, an increase in the density and occupancy of the striatal D2 receptors (D2R) has been frequently reported in schizophrenia patients (Abi-Dargham et al., 2000; Howes and Kapur, 2009; Laruelle et al., 1997; Wong et al., 1986). Many antipsychotic medications primarily aim to block the D2R (la Fougere et al., 2005; Lally and MacCabe, 2015; Yokoi et al., 2002), but the drug dose is the key to the treatment and severer adverse effects are associated with overdose of D2R antagonism (Levine and Ruha, 2012). In addition, prolonged exposure to antipsychotics often causes extrapyramidal symptoms, including Parkinsonian symptoms and tardive dyskinesia (Jarskog et al., 2007; Seeman, 2002). The dose-dependent effects when modulating D2R were also found in cognitive functions such as serial discrimination, in which relatively low and high dose of D2R agonist in striatum impair the performance in the discrimination task, while the intermediate dose of D2R agonist produces significant improvement (Cools and D’Esposito, 2011; Goldman-Rakic et al., 2000; Horst et al., 2019; Mattay et al., 2003). These observations thus further underscore the dynamic interplays and complexity of basal ganglia pathways in action control, as demonstrated in current study and the new triple-control functional model.

Supporting information